---
title: |
  On the Scaling of Volatility with Level in Agricultural Commodity Prices and Yields
type: Letter
author:
  - name: Benjamin Schiek
    affil: a
affiliation:
  - num: a
    address: |
      Alliance of Bioversity International and CIAT, Km 17 Recta Cali-Palmira, Valle del Cauca, Colombia, CP 763537
bibliography: volPropLev.bib
abstract: |
  This note documents a systematic positive association between price level and volatility across agricultural commodities using long-run FAO data. A complementary theoretical result shows that homogeneous production technology implies proportional scaling between yield level and input sensitivity. Together, these observations indicate a smallholder reward-stability tradeoff, underscoring the importance of explicitly accounting for yield and price volatility when formulating  smallholder yield- and income-enhancing research policy.
keywords: |
  Agricultural risk; Commodity price volatility; Yield stability; Fluctuation scaling; Smallholder resilience
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \interfootnotelinepenalty=10000
output: rticles::tf_article
---

```{r setup, include=FALSE, echo=FALSE}
library(tidyverse)
library(patchwork)

knitr::opts_chunk$set(echo = TRUE)

# Plot parameters
shapeVec <- c(21:25, 4)
pointSize <- 1.5
smallPointSize <- 1
labelSize <- 2.5
smallLabelSize <- 2
titleSize <- 8
subtitleSize <- 7
legendTextSize <- 6
axisTextSize <- 6
axisTitleSize <- 7
facetTitleSize <- 7

#Folder
thisFolder <- "D:/OneDrive - CGIAR/Documents 1/CIAT 2/Volatility is proportionate to level in agricultural commodity prices/Data/"

# Functions
processDf <- function(thisFilepath, theseElements, theseUnits, isExpPrice = F){
  dfRaw <- read.csv(thisFilepath, stringsAsFactors = F)
  dfRaw <- dfRaw %>% subset(Element %in% theseElements &
                            Area == "World" & Unit %in% theseUnits) %>%
  select(-contains(".")) %>% select(-matches("[NF]$")) %>% rename_with(~ str_remove(., "^Y"))
  gathercols <- colnames(dfRaw)[-c(1:4)]
  df <- dfRaw %>%
  select(-Unit) %>%
  pivot_longer(cols = all_of(gathercols), names_to = "year", values_to = "val")
  # Keep only items with complete time series
  df <- df %>% group_by(Area, Element, Item) %>% mutate(nYrs = length(year))
  #hist(df$nYrs)
  df <- df %>% subset(nYrs == 64)
  # Keep only items with no zero values in their series
  df <- df %>% group_by(Element, Item) %>% mutate(n0 = sum(val == 0)) %>% subset(n0 == 0)
  if(isExpPrice){
  df <- df %>% pivot_wider(names_from = Element, values_from = val) %>%
        mutate(price = `Export value` / `Export quantity`) %>%
        filter(is.finite(price))
  }
  return(df)
}

plotPriceSeries <- function(thisDfPlotTs, plotTitle, yTitle){
  gg <- ggplot(thisDfPlotTs, aes(x = year, y = priceSeries, color = Item, group = Item))
  gg <- gg + geom_line(lwd = 1)
  gg <- gg + labs(y = yTitle,
                title = plotTitle)
  gg <- gg + theme_bw()
  gg <- gg + theme(axis.title.y = element_text(size = axisTitleSize),
                 axis.text = element_text(size = axisTextSize),
                 legend.text = element_text(size = legendTextSize),
                 plot.title = element_text(size = titleSize),
                 axis.title.x = element_blank(),
                 legend.title = element_blank(),
                 legend.position = "bottom")
  return(gg)
}

processRiskRewardDf <- function(thisDf){
  dfRiskReward <- thisDf %>% group_by(Item) %>% summarise(mPrice = mean(priceSeries),
                                          sPrice = sd(diff(priceSeries)[-1])) %>%
  mutate(lmPrice = log(mPrice), lsPrice = log(sPrice)) %>% select(Item, lmPrice, lsPrice)
  dfRiskReward <- dfRiskReward %>%
  mutate(group = case_when(
    # 1. Animal products
    str_detect(Item, regex("Animal|Dairy|Meat|Bovine|Beef|Pig|Poultry|Sheep|Goat|Milk|Cheese|Butter|Egg|Yoghurt|Whey|Fat of|Honey|Beeswax|Tallow|Ghee|Offal|Lard|Fat|Fatty|Cream|Yoghurt", ignore_case = TRUE)) ~ "Animal products",

    # 2. Cereals & starchy roots
    str_detect(Item, regex("Chicory|Taro|Wheat|Fonio|Rice|Maize|Barley|Oats|Sorghum|Millet|Cereal|Buckwheat|Rye|Grain|Bran|Cassava|Potato|Sweet potatoes|Yams|Roots and Tubers|Starch of|Bread|Pastry|pasta", ignore_case = TRUE)) ~ "Cereals & starchy roots",

    # 3. Fruit & vegetables
    str_detect(Item, regex("Beet pulp|Mangoes|Cucumber|Eggplants (aubergines)|Pumpkins|Lemons|Okra|Papayas|Lettuce|Vegetable|Cherries|Apples|Bananas|Oranges|Grapes|Citrus|Berries|Onions|Tomatoes|Vegetables|Fruit|Juice|Melons|Mushrooms|Asparagus|Cabbages|Pineapples|Plantains|Garlic|Spinach|Peaches|Pears|Apricots|Raisins|Artichoke|Avocado|Cauliflower|Chillies|Carrot|Currants|Dates|Figs|Cider|Plums|Persimmons|Tangerines|Quinces", ignore_case = TRUE)) ~ "Fruit & vegetables",

    # 4. Pulses, nuts, oils
    str_detect(Item, regex("Beans|palm fruit|Peas|Lentils|Chick peas|Pulses|Almonds|Cashew|Walnut|Hazelnut|Nuts|Pistachio|Areca|Oil|Soy|Sunflower|Rape|Palm|Coconut|Olive|Linseed|Sesame|Cake of|Groundnut|Mustard|Castor|Margarine|Copra|Karite|seed", ignore_case = TRUE)) ~ "Pulses, nuts, oils",

    # 5. Sugar & stimulants
    str_detect(Item, regex("Anise|Spice|Hop|Sugar|stimulant|Molasses|Coffee|Tea|Cocoa|Maté|Confectionery|Chocolate|Glucose|Lactose|Alcoholic|Wine|Beer|Beverages|Ethyl alcohol|Cinnamon|Cloves|Ginger|Nutmeg|Malt|Pepper|Vanilla", ignore_case = TRUE)) ~ "Sugar, spices, stimulants",

    # 6. Industrial/non-food
    str_detect(Item, regex("Pyrethrum, dried flowers|Ramie, raw or retted|Sisal, raw|Pellets|Cotton|Wool|Silk|Rubber|Tobacco|Hides|Skins|Abaca|Jute|Coir|Flax|Cigarettes|Cigars|Straw|Hair|Waste|Residues|wax|feed|Hay|forage|materials|Textile|Fibre|cat food", ignore_case = TRUE)) ~ "Industrial/feed/non-food",

    # Default for miscellaneous prep or missed items
    TRUE ~ "Other"
  ))
return(dfRiskReward)
}

getSlopeEtc <- function(dfRiskReward){
  dfMod <- dfRiskReward %>% select(lmPrice, lsPrice) %>% as.data.frame()
row.names(dfMod) <- dfRiskReward$Item
mod <- lm(lmPrice ~., dfMod)
#summary(mod)
b <- round(mod$coefficients[1], 2)
m <- round(mod$coefficients[2], 2)
#plot(mod$fitted.values, mod$residuals)
#---
# gg <- ggplot(data.frame(Item = row.names(dfMod), yhat = mod$fitted.values,
#                         resid = mod$residuals),
#              aes(x = yhat, y = resid, label = Item))
# gg <- gg + geom_point() + geom_text()
# gg <- gg + geom_hline(yintercept = c(-1, 1), color = "red")
#---
# ind_outlier <- which(abs(as.numeric(mod$residuals)) >= 0.75)
# excluded_items <- row.names(dfMod)[ind_outlier]
# pct_removed <- 100 * length(ind_outlier) / nrow(dfRiskReward)
# row_names <- row.names(dfMod)[-ind_outlier]
# dfMod <- dfMod[-ind_outlier, ]
# row.names(dfMod) <- row_names
# mod <- lm(lmPrice ~., dfMod)
# #summary(mod)
# b <- round(mod$coefficients[1], 3)
# m <- round(mod$coefficients[2], 3)
dfOut <- as.data.frame(broom::glance(mod))
adjR2 <- round(dfOut$adj.r.squared, 2)
N <- df.residual(mod)
outVec <- c(slope = m, yInt = b, adjR2 = adjR2, N = N)
return(outVec)
}

```


# Introduction

Modern approaches to agricultural research for development (AR4D) increasingly adopt holistic frameworks, placing strong emphasis on smallholder agronomic and financial stability alongside the conventional focus on genetic gain, high input management, and smallholder integration into high value commodity markets [@conway2007doubly; @altieri2015agroecology; @mockshell2018beyond; @sandlin2014agricultural; @klerkx2010adaptive; @byerlee2009agriculture].

Relatively less attention, however, has been given to whether these objectives are jointly attainable, or whether systematic tradeoffs emerge between reward (i.e., higher yield and/or commodity price) and stability.

This note presents two complementary arguments---one empirical and one theoretical---consistent with the existence of such a tradeoff. First, I document a robust empirical regularity across agricultural commodities: a positive association between price level and price volatility. Second, I show that under standard assumptions of homogeneous production technology, the local sensitivity of yield to input perturbations for a given commodity necessarily scales with yield level.

The note concludes with brief remarks on the implications of these regularities for AR4D policy and research prioritization.


```{r Fig12, fig.show = "hold", fig.width = 6, fig.height=7, fig.align="center", fig.cap="\\label{fig:riskReward}Logged means plotted against logged standard deviations for (A) world export prices, (B) world producer prices per hectare, (C) world producer prices per kg, and (D) yields for agricultural commodities, 1961–2024. Export prices are computed as deflated export values divided by export quantities. Producer prices (per kilogram) are calculated as the FAO Value of Agricultural Production (VAP) divided by production. Producer prices (per hectare) are calculated as VAP divided by area harvested. Standard deviations are computed from first-differenced series. All data are obtained from FAOSTAT (FAO, 2026). Nominal U.S. dollar values are deflated to constant 2014–2016 USD using the World Bank Manufactures Unit Value (MUV) Index (World Bank, 2026).", echo = FALSE, warning=FALSE, message=FALSE}

#============================================================================
# Export price series
#============================================================================
thisFile <- "Trade_CropsLivestock_E_All_Data.csv"
thisFilepath <- paste0(thisFolder, thisFile)
theseElements <- c("Export quantity", "Export value")
theseUnits <- c("t", "1000 USD")
df <- processDf(thisFilepath, theseElements, theseUnits, isExpPrice = T)

# To deflate historical USD values, FAO uses the WB MUV series.
# Let's do the same.
# Available on the "Pink Sheet" (annual) https://www.worldbank.org/en/research/commodity-markets
thisFile <- "CMO-Historical-Data-Annual.xlsx"
thisFilepath <- paste0(thisFolder, thisFile)
dfMUV <- readxl::read_xlsx(thisFilepath, sheet = "Annual Indices (Real)")
dfMUV <- dfMUV %>% select(year = 1, MUV = 18) %>% slice(-c(1:8))
dfMUV$MUV <- as.numeric(dfMUV$MUV)
# Rebase MUV to more suitable year dollars (default is 2010 dollars)
# (Rebase to match FAO 2014-2016 constant USD used in FAO VAP data)
muvRebaser <- dfMUV %>%
  filter(year %in% c("2014", "2015", "2016")) %>%
  summarise(muvRebaser = mean(MUV)) %>%
  pull(muvRebaser)
df <- df %>% merge(dfMUV) %>% mutate(priceSeries = price * muvRebaser / MUV)
# # Get historical CPI deflator series (Historical CPI-U) from Bureao of Labor Statistics
# #https://www.bls.gov/cpi/tables/supplemental-files/
# # Use "Index Averages" worksheet (the annual avg over months)
# # Note this deflates to 1982-84 dollars. To rebase to 2024 dollars: CPIrebased = cpiSeries / (cpi2024value / 100)
# thisFile <- "historical-cpi-u-202512.xlsx"
# thisFilepath <- paste0(thisFolder, thisFile)
# dfCPI <- readxl::read_excel(thisFilepath, sheet = "Index Averages")
# dfCPI <- dfCPI %>% select(year = 2, avgCPI = 5) %>% slice(-c(1:5))
# dfCPI$avgCPI <- as.numeric(dfCPI$avgCPI)
# # Rebase CPI to more suitable year dollars (default is 1982-84 dollars)
# # (Rebase to match FAO 2014-2016 constant USD used in VAP data)
# cpiRebaser <- dfCPI %>%
#   filter(year %in% c("2014", "2015", "2016")) %>%
#   summarise(cpiRebaser = mean(avgCPI)) %>%
#   pull(cpiRebaser)
# df <- df %>% merge(dfCPI) %>% mutate(priceSeries = price * cpiRebaser / avgCPI)
#----------------------------------------------------------------------------
# Check stationarity
# library(Kendall)
# df <- df %>% group_by(Item) %>% mutate(MK_kgPrice = as.numeric(MannKendall(priceSeries)[1]))
# hist(df$MK_kgPrice)
# library(fable)
# library(feasts)
# library(tsibble)
# library(features)
# # 1. Convert your df to a tsibble
# df_ts <- df %>%
#   mutate(year = as.integer(year)) %>%
#   as_tsibble(index = year, key = Item)
# # 2. Calculate features
# df_features <- df_ts %>% features(priceSeries, feat_stl)
# # 3. Look at the 'trend_strength' column
# # This will give you a value 0-1 for every commodity
# head(df_features)
#----------------------------------------------------------------------------
# # Plot a random sample of deflated commodity world export price series
# # num_to_plot <- 5
# # randomItems <- sample(unique(df$Item), num_to_plot)
# #df$Item[grep("Rice", df$Item)] %>% unique()
# plotThese <- c("Wheat", "Rice, milled", "Beans, dry", "Natural Rubber", "Beef and Buffalo Meat, primary", "Sugar Crops Primary")
# yTitle <- "World export real price\n(2014-2016 USD / kg)"
# plotTitle <- "World export real price series\n(1961-2024) for selected commodities"
# dfPlotTs <- df %>% mutate(year = as.integer(year)) %>% subset(Item %in% plotThese)
# gg <- plotPriceSeries(thisDfPlotTs = dfPlotTs, plotTitle, yTitle)
# ggSampleTsExpPrice <- gg
#----------------------------------------------------------------------------
# Risk reward plot
dfRiskReward <- processRiskRewardDf(df)
dfRiskReward <- dfRiskReward %>% subset(group != "Other" & Item != "Crude Materials nes")
# Set group colors to be used in all risk reward plots
n <- length(unique(dfRiskReward$group))
bag_of_colors <- randomcoloR::distinctColorPalette(k = 2 * n)
colorVec <- sample(bag_of_colors, n)
#----------------------------------------------------------------------------
# Get slope and intercept of the risk reward relationship
outVec <- getSlopeEtc(dfRiskReward)
m <- outVec[1]; b <- outVec[2]; adjR2 <- outVec[3]; N <- outVec[4];
thisFacet <- paste0("N = ", N, " agricultural commodities", "\nAdj. R-squared = ", adjR2, ", Slope = ", m, ", Y intercept = ", b)
dfRiskReward$facetKey <- paste0("A) Export price (constant 2014-16 USD/kg)\n", thisFacet)
#----------------------------------------------------------------------------
#dfRiskReward$Item[grep("Rubber", dfRiskReward$Item, ignore.case = T)]
labelItems <- c("Maize (corn)", "Flour of maize",
  #"Wheat", "Flour, wheat",
  #"Starch, cassava",
                "Coffee, green", "Coffee, decaffeinated or roasted", #"Bananas",
                #"Plantains",
                #"Sweet potatoes",
                #"Potatoes",
                "Cocoa beans",
                #"Wool, greasy", 
                #"Beans, dry",
                "Raw silk (not thrown)",
                #"Apples",
                "Vanilla, raw",
                "Soya beans", "Soya bean oil", "Natural Rubber",# "Rice",
                "Cocoa butter, fat and oil")
#---
dfPlot1 <- dfRiskReward
dfPlot1$labelThese <- NA; u <- dfPlot1$Item
dfPlot1$labelThese[which(u %in% labelItems)] <- u[which(u %in% labelItems)]
dfPlot1$group <- as.factor(dfPlot1$group)
names(colorVec) <- levels(dfPlot1$group)
names(shapeVec) <- levels(dfPlot1$group)
fillScale <- scale_fill_manual(name = "group", values = colorVec)
shapeScale <- scale_shape_manual(name = "group", values = shapeVec)

gg <- ggplot(dfPlot1, aes(x = lsPrice, y = lmPrice,
                          group = group,
                          fill = group,
                          shape = group))
gg <- gg + geom_smooth(aes(group = NULL, fill = NULL, shape = NULL), method = lm, se = F)
gg <- gg + geom_point(alpha = 0.6, size = pointSize, color = "black", stroke = 0.5)
gg <- gg + shapeScale
gg <- gg + fillScale
gg <- gg + labs(x = "Nat. log of stand. dev.", y = "Nat. log of mean")
gg <- gg + ggrepel::geom_text_repel(aes(label = labelThese), size = labelSize,
                           point.padding = 1,
                           box.padding = 1,
                           nudge_x = 1,
                           nudge_y = 1,
                           max.overlaps = Inf)
gg <- gg + facet_wrap(~facetKey)
gg <- gg + theme_bw()
gg <- gg + theme(axis.title = element_text(size = axisTitleSize),
                 axis.text = element_text(size = axisTextSize),
                 legend.position = "none",
                 strip.background = element_rect(fill = "white"),
                 strip.text = element_text(hjust = 0, size = axisTitleSize))
gg1 <- gg
#============================================================================
# Producer price and yield series
#============================================================================
thisFile <- "Production_Crops_Livestock_E_All_Data.csv"
thisFilepath <- paste0(thisFolder, thisFile)
theseElements <- c("Production", "Area harvested", "Yield"); theseUnits <- c("t", "ha", "kg/ha")
dfAreaProd <- processDf(thisFilepath, theseElements, theseUnits, isExpPrice = F) %>%
  spread(Element, val) %>% select(Item, year, Production, `Area harvested`, Yield)
thisFile <- "Value_of_Production_E_All_Data.csv"
thisFilepath <- paste0(thisFolder, thisFile)
theseElements <- "Gross Production Value (constant 2014-2016 thousand US$)"; theseUnits <- "1000 USD"
dfVAP <- processDf(thisFilepath, theseElements, theseUnits, isExpPrice = F) %>%
  select(Item, year, val) %>% rename(VAP = val); dfVAP$Element <- NULL
df <- dfAreaProd %>% merge(dfVAP, by = c("Item", "year")) %>%
  mutate(haPrice = 1000 * VAP / `Area harvested`, kgPrice = VAP / Production,
         year = as.integer(year))
#----------------------------------------------------------------------------
# Check stationarity
# library(Kendall)
# # Example for one item:
# df <- df %>% group_by(Item) %>% mutate(MK_kgPrice = as.numeric(MannKendall(kgPrice)[1]), MK_haPrice = as.numeric(MannKendall(haPrice)[1]))
# hist(df$MK_kgPrice)
# hist(df$MK_haPrice)
# x <- MannKendall(df$kgPrice[df$Item == "Abaca, manila hemp, raw"])
#----------------------------------------------------------------------------
# # Plot a random sample of deflated commodity world price series
# #df$Item[grep("Soya beans", df$Item, ignore.case = T)] %>% unique()
# #"Coffee, green", "Wheat", "Soya beans",
# plotThese <- c("Sugar Crops Primary", "Walnuts, in shell",
#                "Pistachios, in shell", "Beeswax",
#               "Raspberries", "Sesame seed")
# # num_to_plot <- 15
# # randomItems <- sample(unique(df$Item), num_to_plot)
# # plotThese <- randomItems
# yTitle <-"World producer real price\n(2014-2016 USD / kg)"
# plotTitle <- "World producer real price series\n(1961-2024) for selected commodities (constant 2014-2016 USD)"
# dfPlotTs <- df %>% subset(Item %in% plotThese) %>% rename(priceSeries = kgPrice)
# gg <- plotPriceSeries(dfPlotTs, plotTitle, yTitle)
# ggSampleTsProdPrice <- gg
# yTitle <-"World producer real price per hectare\n(2014-2016 USD / hectare)"
# plotTitle <- "World producer real production value per hectare series\n(1961-2024) for selected commodities (constant 2014-2016 USD)"
# dfPlotTs <- df %>% subset(Item %in% plotThese) %>% rename(priceSeries = haPrice)
# gg <- plotPriceSeries(dfPlotTs, plotTitle, yTitle)
# ggSampleTsProdHaPrice <- gg
#----------------------------------------------------------------------------
# Risk reward plot
dfKg <- df %>% rename(priceSeries = kgPrice)
dfRiskReward <- processRiskRewardDf(dfKg)
dfRiskReward <- dfRiskReward %>% subset(group != "Other")
#----------------------------------------------------------------------------
# Get slope and intercept of the risk reward relationship
outVec <- getSlopeEtc(dfRiskReward)
m <- outVec[1]; b <- outVec[2]; adjR2 <- outVec[3]; N <- outVec[4];
dfKgPrice <- dfRiskReward
thisFacet <- paste0("N = ", N, " agricultural commodities", "\nAdj. R-squared = ", adjR2, ", Slope = ", m, ", Y intercept = ", b)
dfKgPrice$facetKey <- paste0("C) Producer price (constant 2014-16 USD/kg)\n", thisFacet)
#----------------------------------------------------------------------------
# labelItems <- c("Maize (corn)",
#                 "Coffee, green",
#                 "Bananas",
#                 #"Plantains",
#                 #"Sweet potatoes",
#                 #"Potatoes",
#                 "Cocoa beans",
#                 #"Wool, greasy", 
#                 "Beans, dry",
#                 "Silk-worm cocoons suitable for reeling",
#                 #"Apples",
#                 "Vanilla, raw",
#                 "Soya beans", "Natural rubber in primary forms", "Rice")
#----------------------------------------------------------------------------
dfYd <- df %>% rename(priceSeries = Yield)
dfRiskReward <- processRiskRewardDf(dfYd)
dfRiskReward <- dfRiskReward %>% subset(group != "Other")
#----------------------------------------------------------------------------
# Get slope and intercept of the risk reward relationship
outVec <- getSlopeEtc(dfRiskReward)
m <- outVec[1]; b <- outVec[2]; adjR2 <- outVec[3]; N <- outVec[4];
dfYd <- dfRiskReward
thisFacet <- paste0("N = ", N, " agricultural commodities", "\nAdj. R-squared = ", adjR2, ", Slope = ", m, ", Y intercept = ", b)
dfYd$facetKey <- paste0("D) Yield (kg/ha)\n", thisFacet)
#----------------------------------------------------------------------------
dfHa <- df %>% rename(priceSeries = haPrice)
dfRiskReward <- processRiskRewardDf(dfHa)
dfRiskReward <- dfRiskReward %>% subset(group != "Other")
#----------------------------------------------------------------------------
# Get slope and intercept of the risk reward relationship
outVec <- getSlopeEtc(dfRiskReward)
m <- outVec[1]; b <- outVec[2]; adjR2 <- outVec[3]; N <- outVec[4];
dfHaPrice <- dfRiskReward
thisFacet <- paste0("N = ", N, " agricultural commodities", "\nAdj. R-squared = ", adjR2, ", Slope = ", m, ", Y intercept = ", b)
dfHaPrice$facetKey <- paste0("B) Producer price (constant 2014-16 USD/hectare)\n", thisFacet)
#----------------------------------------------------------------------------
dfPlot2 <- do.call(rbind, list(dfKgPrice, dfHaPrice, dfYd)) %>% as.data.frame()
dfPlot2$group <- as.factor(dfPlot2$group)
names(colorVec) <- levels(dfPlot2$group)
names(shapeVec) <- levels(dfPlot2$group)
fillScale <- scale_fill_manual(name = "group", values = colorVec)
shapeScale <- scale_shape_manual(name = "group", values = shapeVec)

gg <- ggplot(dfPlot2, aes(x = lsPrice, y = lmPrice,
                          group = group,
                          fill = group,
                          shape = group))
gg <- gg + geom_smooth(aes(group = NULL, fill = NULL, shape = NULL), method = lm, se = F)
gg <- gg + geom_point(alpha = 0.6, size = pointSize, color = "black", stroke = 0.5)
gg <- gg + shapeScale
gg <- gg + fillScale
gg <- gg + labs(x = "Nat. log of stand. dev.", y = "Nat. log of mean")
gg <- gg + facet_wrap(~facetKey, nrow = 2, scales = "free")
gg <- gg + theme_bw()
gg <- gg + theme(axis.title = element_text(size = axisTitleSize),
                 axis.text = element_text(size = axisTextSize),
                 legend.position = "right",
                 legend.spacing.x = unit(0.25, 'cm'),
                 legend.text = element_text(size = legendTextSize),
                 legend.title = element_blank(),
                 strip.background = element_rect(fill = "white"),
                 strip.text = element_text(hjust = 0, size = axisTitleSize))
gg <- gg + guides(
  fill = guide_legend(
    ncol = 2,               # Set to 2 columns
    byrow = TRUE, 
    override.aes = list(linetype = 0, shape = shapeVec) # Ensure fill shows correctly
  ),
  shape = guide_legend(
    ncol = 2                # Keep columns consistent across scales
  ),
  color = guide_legend(
    ncol = 2,
    override.aes = list(linetype = 0)
  )
)
gg <- gg + theme(
  legend.position = c(0.8, 0.2), # Adjust coordinates to land in the empty facet
  legend.justification = c(0.6, 0.5)
)
gg2 <- gg

(gg1 / gg2) + 
  plot_annotation(
    title = "Empirical evidence of a reward-stability tradeoff in commodity prices and yields",
    subtitle = "FAO price and yield series 1961-2024, dollar values in constant 2014-16 USD",
    theme = theme(
      plot.title = element_text(size = titleSize),
      plot.subtitle = element_text(size = subtitleSize)
    )
  )

```

# Empirical evidence of a reward-stability tradeoff in agricultural commodities

To assess the relationship between agricultural risk and reward, the mean and standard deviation of global yield and price series over 1961-2024, computed from FAO data [-@FAOSTAT2026], are plotted against each other in logs (Figure \ref{fig:riskReward}). Only commodities with a complete, non-zero time series are included. Comparisons of moments across time series are meaningful insofar as the series are stationary. To ensure stationarity, standard deviations are computed from first-differenced series. In the price series (panels A-C), nominal values are deflated using the World Bank Manufacturers Unit Value Index [@WB_MUV_2026], such that the means are expressed in constant 2014-16 US dollars. The means of yield series (panel D), however, should be interpreted with great caution, since yield series generally exhibit a secular trend. Producer price is displayed in both per unit weight and per unit area harvested terms, indicating that the observed regularity is robust to alternative value denominations.

The plots reveal a strong empirical correspondence between level and volatility across a wide range of commodity types, including crop, animal, food, non-food, raw, and processed commodities. The formal economic mechanism behind the regularity, if any exists, does not readily suggest itself. Similar scaling relations have often been observed in other contexts---primarily biological and ecological systems---where they are known as Taylor’s Law or fluctuation scaling [@eisler2008fluctuation]. However, the explanations offered in those contexts are domain specific and do not readily extend to agricultural commodity markets. The search for abstract statistical explanations not rooted in any particular empirical context is an active area of research [@cohen2015random; @brown2021taylor; @taylor2019taylor].
<!-- Notably, the scaling relation appears consistently across commodities with markedly different biological, technological, and market characteristics. -->
[-@WB_MUV_2026;]

# Theoretical basis for a yield level-stability tradeoff

At the farm level, it is straightforward to demonstrate a fundamental theoretical reason to expect a tradeoff between yield level and yield stability.

Under the standard assumption that yield $y(x)$ is a homogeneous function of inputs $x$ of degree $h$, Euler’s theorem implies

\begin{equation}
\sum_{i=1}^n x_i \frac{\partial y}{\partial x_i}
= h y(\mathbf{x})
\end{equation}

or equivalently,

\begin{equation}
\mathbf{x}' \nabla y(\mathbf{x})
= h y(\mathbf{x}).
\end{equation}

Differentiating both sides yields

\begin{equation}
\nabla \left( \mathbf{x}' \nabla y \right)
= \nabla (h y).
\end{equation}

Applying the product rule,

\begin{equation}
\nabla y + H(\mathbf{x}) \mathbf{x}
= h \nabla y,
\end{equation}

where $H(x)$ denotes the Hessian matrix.

Rearranging,

\begin{equation}
H(\mathbf{x}) \mathbf{x}
= (h - 1)\nabla y.
\end{equation}

Premultiplying by $x'$

\begin{equation}
\mathbf{x}' H(\mathbf{x}) \mathbf{x}
= (h - 1) \mathbf{x}' \nabla y.
\end{equation}

Invoking Euler’s theorem once more,

\begin{equation}
\mathbf{x}' H(\mathbf{x}) \mathbf{x}
= h(h - 1) y(\mathbf{x}).
\label{eq:volatPropToLev}
\end{equation}

The quadratic form $\mathbf{x}'H(\mathbf{x}) \mathbf{x}$ captures local curvature of the yield function and therefore measures sensitivity of output to small perturbations of inputs.

For a strictly concave production technology, the Hessian is negative definite, implying

\begin{equation}
\mathbf{x}' H(\mathbf{x}) \mathbf{x} < 0.
\end{equation}

This condition is satisfied when $0 < h < 1$, corresponding to decreasing returns to scale. Equation \ref{eq:volatPropToLev} therefore implies that curvature---and hence local input sensitivity---scales proportionately with yield level.

In economic terms, higher-yield production regimes are inherently more sensitive to deviations of inputs from their optimal values, providing a theoretical basis for a yield level–stability tradeoff.

# Discussion and conclusion

The empirical and theoretical results presented above should be clearly distinguished. The theoretical argument demonstrates that yield sensitivity to stochastic disturbances (e.g., rainfall deviations) necessarily scales with yield level under standard assumptions. Farm-level yield volatility therefore reflects both stochastic shocks and the scale-dependent response of output to those shocks.
 <!-- Yield volatility is thus attributable not only to exogenous disturbances, but also to the intrinsic sensitivity to distrubances, which increases with yield level. -->

The empirical regularity, by contrast, indicates that long-run agricultural commodity price volatility scales with price level, although the mechanism underlying this relationship remains unclear.^[The empirical results are based on world-level aggregates. Future work may examine the extent to which these regularities persist under geographical disaggregation, where deviations or slope heterogeneity may offer additional insights.] The observed scaling may arise from scale-dependent meteorological/agronomic/market disturbances, scale-dependent price responses to these disturbances, or some combination thereof. While the theoretical results establish scale-dependent volatility in production at the farm level, no direct causal link between yield volatility scaling and price volatility scaling is asserted here. Establishing such a connection remains an open question for future research.

From an AR4D policy perspective, the results suggest that efforts to increase farm yield and income are unlikely to be separable from considerations of stability. Productivity and volatility appear intrinsically linked through scale-dependent mechanisms, implying that interventions aiming for higher returns may simultaneously increase exposure to risk.

This underscores the importance of risk-mitigating mechanisms---including financial instruments, institutional arrangements, social capital, and cooperative risk-sharing structures---as central components rather than peripheral complements to yield- and income-enhancing strategies. Policies aimed at productivity growth may therefore benefit from explicit assessment of how volatility scales with expected returns, together with an evaluation of risk-bearing capacity within target populations and production environments.

Contemporary AR4D frameworks (e.g., "Agricultural Innovation Systems" [@sandlin2014agricultural] and "Blended Sustainability" [@mockshell2018beyond]) are already philosophically oriented in this direction. The empirical and theoretical observations presented here provide further quantitative grounding to facilitate the translation of the vision into concrete practice.
<!-- The empirical and theoretical results presented above should be carefully distinguished from each other. The theoretical argument demonstrates that yield sensitivity to random shocks (e.g., low rainfall) on any given farm necessarily scales with yield level under standard assumptions. This means that farm level volatility is partly random and partly deterministic. It is attributable not only to random weather events that are independent of yield level, but also to yield sensitivity to said random events, which scales with yield level. -->
<!-- The empirical regularity, on the other hand, indicates that long run agricultural commodity price volatility scales with price level; but the mechanism behind this scaling relation is unclear.^[Note that the empirical results are based on world-level aggregates. Future work may examine the extent to which these regularities persist under geographical disaggregation, where deviations or slope heterogeneity may offer additional insights.] It may be because random exogenous price shocks themselves scale with price level (for some unknown or speculative reason), or because price sensitivity to exogenous shocks scales with level, or because of some combination thereof, or because of something else entirely. Some may be tempted to draw a connection between the theoretically proven scale-dependent farm yield volatility, and the empirically observed scale-dependent price volatility; but deeper analysis is required to conclusively demonstrate such a connection. -->

<!-- From an AR4D policy perspective, the results presented here imply that higher farm yield and income cannot be pursued independently of higher stability. Rather, the two objectives are deeply intertwined, and inversely proportionate to each other. It follows that any planned progress towards one objective must include a strategy to mitigate regress as regards the other strategy. Concretely, it implies that smallholder risk mitigating measures---microfinance, insurance, subsidies, risk-pooling cooperative mechanisms, other forms of social capital, etc.---are not optional complements to, but rather critical central components of, any smallholder yield and farm income enhancing research strategy, without which the strategy is unlikely to succeed. This, in turn, implies that AR4D strategy should start from a precise quantitative assessment of the increased agronomic and price risk that inevitably accompanies higher yields and exposure to commodity markets---based on the empirical and theoretical scaling laws outlined above---together with an assessment of risk-bearing capacity within target populations and production environments. A precisely "right sized" mitigation strategy, tailored to the particular cultural, socioeconomic, and ecological circumstances of the beneficiary communities may then be planned by comparing assessed risk against risk-bearing capacity. -->

<!-- Modern AR4D strategic visions such as "Agricultural Innovation Systems" [@sandlin2014agricultural] or "Blended Sustainability" [@mockshell2018beyond] are already philosophically oriented in this direction. The present note provides further empirical and theoretical grounding for this vision. -->
<!-- the pursuit of higher yields and greater participation in high-value commodity markets may conflict with objectives centered on yield stability and smallholder resilience. -->
<!-- In this context, yield-enhancing technologies and practices may be most effective when complemented by mechanisms that mitigate exposure to variability, including financial instruments, institutional arrangements, and cooperative risk-pooling structures. Frameworks emphasizing joint productivity and stability objectives may thus benefit from a more explicit recognition of scale-dependent variability. -->
<!-- The empirical regularity and theoretical argument outlined above indicate that agricultural yield and price volatility are not due to random exogenous shocks alone, but rather arise in a partly deterministic/endogenous/mechanical [which word to use?] manner from scale-dependent sensitivity to random shocks---resulting in the observed reward-stability tradeoff. It is important to keep in mind that the empirical findings are world level. A more refined analysis can examine to what degree the regularity persists or whether anything interesting happens (e.g. changing slope) under geographical disaggregation. -->
<!-- For AR4D policy purposes, this means that the pursuit of increased yields and smallholder exposure to high value commodity markets may work against the simultaneous pursuit of higher yield stability and overall smallholder resilience. AR4D yield enhancement strategy should thus start from a careful assessment of the risk capital available in target populations and environments; and complement yield enhancing technology and practices accordingly with the development of appropriate risk mitigating mechanisms (e.g. microfinance, subsidies, insurance, cooperative social mechanisms). Modern AR4D frameworks emphasizing both higher yields and higher stability such as the "Agricultural Innovation Systems" framework [@sandlin2014agricultural; @klerkx2010adaptive] or the "Blended Sustainability" vision [@mockshell2018beyond] can thus take a more grounded, informed approach. -->
<!-- Above I document an empirical tradeoff between stability and reward in world commodity markets, and demonstrate theoretically that, at farm level, yield volatility scales with yield level under standard microeconomic assumptions regarding homogeneity and concavity. This strongly -->

