---
title: |
  On the Scaling of Volatility with Level in Agricultural Commodity Prices and Yields
type: ARTICLE TEMPLATE
author:
  - name: Benjamin Schiek
    affil: a, \dagger
affiliation:
  - num: a
    address: |
      Alliance of Bioversity International and CIAT, Km 17 Recta Cali-Palmira, Valle del Cauca, Colombia, CP 763537
bibliography: volPropLev.bib
abstract: |
  This note documents a systematic positive association between price level and volatility across agricultural commodities using long-run FAO data. A complementary theoretical result shows that homogeneous production technology implies proportional scaling between yield level and input sensitivity. The findings highlight the role of scale in shaping agricultural variability.
keywords: |
  Sections; lists; figures; tables; mathematics; fonts; references; appendices
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
---

```{r setup, include=FALSE}
library(tidyverse)
library(patchwork)

knitr::opts_chunk$set(echo = TRUE)

# Plot parameters
shapeVec <- c(21:25, 4)
pointSize <- 1.5
smallPointSize <- 1
labelSize <- 2.5
smallLabelSize <- 2
titleSize <- 8
subtitleSize <- 7
legendTextSize <- 7
axisTextSize <- 6
axisTitleSize <- 7
facetTitleSize <- 7

#Folder
thisFolder <- "D:/OneDrive - CGIAR/Documents 1/CIAT 2/Volatility is proportionate to level in agricultural commodity prices/Data/"

# Functions
processDf <- function(thisFilepath, theseElements, theseUnits, isExpPrice = F){
  dfRaw <- read.csv(thisFilepath, stringsAsFactors = F)
  dfRaw <- dfRaw %>% subset(Element %in% theseElements &
                            Area == "World" & Unit %in% theseUnits) %>%
  select(-contains(".")) %>% select(-matches("[NF]$")) %>% rename_with(~ str_remove(., "^Y"))
  gathercols <- colnames(dfRaw)[-c(1:4)]
  df <- dfRaw %>%
  select(-Unit) %>%
  pivot_longer(cols = all_of(gathercols), names_to = "year", values_to = "val")
  # Keep only items with complete time series
  df <- df %>% group_by(Area, Element, Item) %>% mutate(nYrs = length(year))
  #hist(df$nYrs)
  df <- df %>% subset(nYrs == 64)
  # Keep only items with no zero values in their series
  df <- df %>% group_by(Element, Item) %>% mutate(n0 = sum(val == 0)) %>% subset(n0 == 0)
  if(isExpPrice){
  df <- df %>% pivot_wider(names_from = Element, values_from = val) %>%
        mutate(price = `Export value` / `Export quantity`) %>%
        filter(is.finite(price))
  }
  return(df)
}

plotPriceSeries <- function(thisDfPlotTs, plotTitle, yTitle){
  gg <- ggplot(thisDfPlotTs, aes(x = year, y = priceSeries, color = Item, group = Item))
  gg <- gg + geom_line(lwd = 1)
  gg <- gg + labs(y = yTitle,
                title = plotTitle)
  gg <- gg + theme_bw()
  gg <- gg + theme(axis.title.y = element_text(size = axisTitleSize),
                 axis.text = element_text(size = axisTextSize),
                 legend.text = element_text(size = legendTextSize),
                 plot.title = element_text(size = titleSize),
                 axis.title.x = element_blank(),
                 legend.title = element_blank(),
                 legend.position = "bottom")
  return(gg)
}

processRiskRewardDf <- function(thisDf){
  dfRiskReward <- thisDf %>% group_by(Item) %>% summarise(mPrice = mean(priceSeries),
                                          sPrice = sd(diff(priceSeries)[-1])) %>%
  mutate(lmPrice = log(mPrice), lsPrice = log(sPrice)) %>% select(Item, lmPrice, lsPrice)
  dfRiskReward <- dfRiskReward %>%
  mutate(group = case_when(
    # 1. Animal products
    str_detect(Item, regex("Animal|Dairy|Meat|Bovine|Beef|Pig|Poultry|Sheep|Goat|Milk|Cheese|Butter|Egg|Yoghurt|Whey|Fat of|Honey|Beeswax|Tallow|Ghee|Offal|Lard|Fat|Fatty|Cream|Yoghurt", ignore_case = TRUE)) ~ "Animal products",

    # 2. Cereals & starchy roots
    str_detect(Item, regex("Chicory|Taro|Wheat|Fonio|Rice|Maize|Barley|Oats|Sorghum|Millet|Cereal|Buckwheat|Rye|Grain|Bran|Cassava|Potato|Sweet potatoes|Yams|Roots and Tubers|Starch of|Bread|Pastry|pasta", ignore_case = TRUE)) ~ "Cereals & starchy roots",

    # 3. Fruit & vegetables
    str_detect(Item, regex("Beet pulp|Mangoes|Cucumber|Eggplants (aubergines)|Pumpkins|Lemons|Okra|Papayas|Lettuce|Vegetable|Cherries|Apples|Bananas|Oranges|Grapes|Citrus|Berries|Onions|Tomatoes|Vegetables|Fruit|Juice|Melons|Mushrooms|Asparagus|Cabbages|Pineapples|Plantains|Garlic|Spinach|Peaches|Pears|Apricots|Raisins|Artichoke|Avocado|Cauliflower|Chillies|Carrot|Currants|Dates|Figs|Cider|Plums|Persimmons|Tangerines|Quinces", ignore_case = TRUE)) ~ "Fruit & vegetables",

    # 4. Pulses, nuts, oils
    str_detect(Item, regex("Beans|palm fruit|Peas|Lentils|Chick peas|Pulses|Almonds|Cashew|Walnut|Hazelnut|Nuts|Pistachio|Areca|Oil|Soy|Sunflower|Rape|Palm|Coconut|Olive|Linseed|Sesame|Cake of|Groundnut|Mustard|Castor|Margarine|Copra|Karite|seed", ignore_case = TRUE)) ~ "Pulses, nuts, oils",

    # 5. Sugar & stimulants
    str_detect(Item, regex("Anise|Spice|Hop|Sugar|stimulant|Molasses|Coffee|Tea|Cocoa|Maté|Confectionery|Chocolate|Glucose|Lactose|Alcoholic|Wine|Beer|Beverages|Ethyl alcohol|Cinnamon|Cloves|Ginger|Nutmeg|Malt|Pepper|Vanilla", ignore_case = TRUE)) ~ "Sugar, spices, stimulants",

    # 6. Industrial/non-food
    str_detect(Item, regex("Pyrethrum, dried flowers|Ramie, raw or retted|Sisal, raw|Pellets|Cotton|Wool|Silk|Rubber|Tobacco|Hides|Skins|Abaca|Jute|Coir|Flax|Cigarettes|Cigars|Straw|Hair|Waste|Residues|wax|feed|Hay|forage|materials|Textile|Fibre|cat food", ignore_case = TRUE)) ~ "Industrial/feed/non-food",

    # Default for miscellaneous prep or missed items
    TRUE ~ "Other"
  ))
return(dfRiskReward)
}

getSlopeEtc <- function(dfRiskReward){
  dfMod <- dfRiskReward %>% select(lmPrice, lsPrice) %>% as.data.frame()
row.names(dfMod) <- dfRiskReward$Item
mod <- lm(lmPrice ~., dfMod)
#summary(mod)
b <- round(mod$coefficients[1], 2)
m <- round(mod$coefficients[2], 2)
#plot(mod$fitted.values, mod$residuals)
#---
# gg <- ggplot(data.frame(Item = row.names(dfMod), yhat = mod$fitted.values,
#                         resid = mod$residuals),
#              aes(x = yhat, y = resid, label = Item))
# gg <- gg + geom_point() + geom_text()
# gg <- gg + geom_hline(yintercept = c(-1, 1), color = "red")
#---
# ind_outlier <- which(abs(as.numeric(mod$residuals)) >= 0.75)
# excluded_items <- row.names(dfMod)[ind_outlier]
# pct_removed <- 100 * length(ind_outlier) / nrow(dfRiskReward)
# row_names <- row.names(dfMod)[-ind_outlier]
# dfMod <- dfMod[-ind_outlier, ]
# row.names(dfMod) <- row_names
# mod <- lm(lmPrice ~., dfMod)
# #summary(mod)
# b <- round(mod$coefficients[1], 3)
# m <- round(mod$coefficients[2], 3)
dfOut <- as.data.frame(broom::glance(mod))
adjR2 <- round(dfOut$adj.r.squared, 2)
N <- df.residual(mod)
outVec <- c(slope = m, yInt = b, adjR2 = adjR2, N = N)
return(outVec)
}


plotRiskReward <- function(dfPlot, labelItems, thisTitle = NULL, thisXlab = NULL, thisYlab = NULL, legendOn = F, colorVec, m, b, N, adjR2){
  dfPlot$labelThese <- NA
  u <- dfPlot$Item
  dfPlot$labelThese[which(u %in% labelItems)] <- u[which(u %in% labelItems)]
#---
  thisSubtitle <- paste0("N = ", N, " agricultural commodities", "\nAdj. R-squared = ", adjR2, ", Slope = ", m, ", Y intercept = ", b)
#---
thisTitle <- paste0(thisTitle,"\n", thisSubtitle)
#names(facetLabel) <- facetKey
#colnames(dfPlot)[2:3] <- c("Logged mean", "Logged stand. dev.")
#---
dfPlot$group <- as.factor(dfPlot$group)
names(colorVec) <- levels(dfPlot$group)
names(shapeVec) <- levels(dfPlot$group)
fillScale <- scale_fill_manual(name = "group", values = colorVec)
shapeScale <- scale_shape_manual(name = "group", values = shapeVec)
#---
gg <- ggplot(dfPlot,
             aes(x = lmPrice,
                 y = lsPrice,
                 group = group, fill = group,
                 shape = group,
                 label = labelThese))
gg <- gg + geom_smooth(aes(group = NULL, fill = NULL, shape = NULL), method = lm, se = F)
gg <- gg + geom_point(alpha = 0.6, size = pointSize, color = "black", stroke = 0.5)#,
#fill = "khaki1", pch=21, color="Black")
#gg <- gg + scale_fill_manual(values = color_vec)
# gg <- gg + scale_shape_manual(values = shape_vec)
gg <- gg + shapeScale
gg <- gg + fillScale
gg <- gg + ggrepel::geom_text_repel(size = labelSize,
                           point.padding = 1,
                           box.padding = 1,
                           nudge_x = 1,
                           nudge_y = 1)
#gg <- gg + facet_wrap(~Element, labeller = labeller(Element = facetLabel))
#gg <- gg + labs(title = "The ag commodity producer's risk-reward frontier",
                # subtitle = this_subtitle)
gg <- gg + guides(fill = guide_legend(nrow = 1, byrow = T, override.aes = list(linetype = 0)),
                  color = guide_legend(override.aes = list(linetype = 0)))
gg <- gg + labs(title = thisTitle)
if(!is.null(thisXlab)){
  gg <- gg + labs(x = thisXlab)
  gg <- gg + theme(axis.title.x = element_text(size = axisTitleSize))
}else{
  gg <- gg + theme(axis.title.x = element_blank())
}
if(!is.null(thisYlab)){
  gg <- gg + labs(y = thisYlab)
  gg <- gg + theme(axis.title.y = element_text(size = axisTitleSize))
}else{
  gg <- gg + theme(axis.title.y = element_blank())
}
gg <- gg + theme(plot.title = element_text(size = titleSize),
                 plot.subtitle = element_text(size = subtitleSize),
                 axis.title = element_text(size = axisTitleSize),
                 axis.text = element_text(size = axisTextSize))
if(legendOn){
  gg <- gg + theme(legend.position = "bottom",
                 legend.spacing.x = unit(0.25, 'cm'),
                 legend.text = element_text(size = legendTextSize),
                 legend.title = element_blank())
}else{
  gg <- gg + theme(legend.position = "none")
}
return(gg)
  
}



```

<!-- This note documents a systematic positive association between average prices and price volatility across agricultural commodities using long-run FAO data. The observed pattern is robust and consistent with a simple scaling regularity linking level and variability. A complementary theoretical result shows that, under standard homogeneity properties of production technology, yield level is intrinsically linked to local input sensitivity. Together, these perspectives highlight the role of scale in shaping agricultural variability. -->
<!-- This note documents a strong positive relationship between average prices and price volatility across agricultural commodities using long-run FAO data. The pattern is robust and suggests a simple scaling relationship between level and volatility. I then provide a brief theoretical argument showing that an analogous scaling between yield level and volatility is consistent with basic homogeneity properties of production technology, under which sensitivity to shocks increases with output. The results highlight the importance of scale in shaping volatility and have implications for how risk is conceptualized in agricultural production. -->

# Introduction

Modern approaches to agricultural research for development (AR4D) increasingly adopt holistic frameworks, placing greater emphasis on smallholder agronomic and financial stability alongside the conventional focus on genetic gain, high input management, and smallholder integration into high value commodity markets [@conway2007doubly; @altieri2015agroecology; @mockshell2018beyond; @sandlin2014agricultural; @klerkx2010adaptive].

Relatively less attention, however, has been given to whether these objectives are jointly attainable, or whether systematic tradeoffs emerge between reward (i.e., higher yield or commodity price) and stability.

This note presents two complementary arguments---one empirical and one theoretical---consistent with the existence of such a tradeoff. First, I document a robust empirical regularity across agricultural commodities: a positive association between level and volatility. Second, I show that under standard assumptions of homogeneous production technology, the local sensitivity of yield to input perturbations for a given commodity necessarily scales with yield level.

The note concludes with brief remarks on the implications of these regularities for AR4D policy and research prioritization.

<!-- Modern approaches to agricultural research for development (AR4D) are increasingly holistic, placing strong emphasis on smallholder agronomic and financial stability as a primary research objective to be pursued alongside the more conventional focus on genetic gain and high input, yield enhancing agronomic practices [@conway2007doubly; @altieri2015agroecology; @mockshell2018beyond; @sandlin2014agricultural; @klerkx2010adaptive]. However, little consideration is given to whether it is possible to pursue both higher stability and higher reward (i.e., higher yield and price) simultaneously; or whether a tradeoff exists, such that progress towards one objective nullifies progress toward the other. -->
<!-- In this note, I present two complementary arguments---one empirical and the other theoretical---strongly indicative of a smallholder reward-stability tradeoff. First, I document a robust empirical regularity across agricultural commodities: the higher the value of a commodity, the higher its volatility. Second, I present a brief theoretical argument showing that, under the standard assumptions of homogeneous production technology, yield sensitivity to input perturbations for any single given commodity necessarily scales with yield level. -->
<!-- I conclude with brief remarks on the AR4D policy implications of these tradeoffs. -->
<!-- Modern conventional agricultural research for development (AR4D) narratives and agroecological counter-narratives alike agree on the importance of yield stability and price risk in smallholder planting decisions and welfare outcomes [@conway2007doubly; @altieri2015agroecology]. That is to say, it is generally accepted that the primary objectives of AR4D extend beyond genetic gain and smallholder participation in high value commodity markets to also encompass agrononomic and financial stability. However, little consideration has been given to what extent it is possible to pursue the objectives of higher value and stability simultaneously, or whether a stability-reward tradeoff exists, such that the pursuit of more reward (higher yields and price) necessarily entails higher risk (yield and price volatility). -->
<!-- a cursory but rigorous exploration of this question based on FAO data. The result of the exploration indicates a strong empirical correspondence between yield risk and level, and also between price risk and level.  -->
<!-- Likewise, the higher a commodity yield volatility is likewise proportionate to yield level, indicating a risk–reward tradeoff at the commodity level. -->
<!-- Taken together, these results suggest that yield and price volatility are not merely exogenous uncertainty, but are structurally inherent to scale itself. AR4D projects aiming to raise yields and increase smallholder participation in high value commodity markets should complement with mechanisms   resilience building  -->
<!-- Strategies to get smallholders to adopt high yielding varieties must complement the technology with resilience building cooperative social mechanisms risk pooling. -->

<!-- Means plotted against standard deviations of (panel A) world export price, (panel B) world producer kilogram price, (panel C) world producer hectare price, and (panel D) yield, for agricultural commodities, 1961-2024. Export price = Deflated export value / export quantity. Producer kg price = FAO Value of Agricultural Production (VAP) series divided / Production. Producer ha price = VAP series divided by area harvested. All data are downloaded from FAO (FAO 2026). Nominal dollar values are deflated to constant 2014-16 USD using the World Bank Manufacturers Unit Value Index (World Bank 2026). -->

```{r Fig12, fig.show = "hold", fig.width = 7, fig.height=7, fig.align="center", fig.cap="\\label{fig:riskReward}Means plotted against standard deviations for (A) world export prices, (B) world producer prices (per kilogram), (C) world producer prices (per hectare), and (D) yields for agricultural commodities, 1961–2024. Axes are shown in logarithmic scale. Export prices are computed as deflated export values divided by export quantities. Producer prices (per kilogram) are calculated as the FAO Value of Agricultural Production (VAP) divided by production. Producer prices (per hectare) are calculated as VAP divided by area harvested. Standard deviations are computed from first-differenced series. All data are obtained from FAOSTAT (FAO, 2026). Nominal U.S. dollar values are deflated to constant 2014–2016 USD using the World Bank Manufactures Unit Value (MUV) Index (World Bank, 2026).", echo = FALSE}

#============================================================================
# Export price series
#============================================================================
thisFile <- "Trade_CropsLivestock_E_All_Data.csv"
thisFilepath <- paste0(thisFolder, thisFile)
theseElements <- c("Export quantity", "Export value")
theseUnits <- c("t", "1000 USD")
df <- processDf(thisFilepath, theseElements, theseUnits, isExpPrice = T)

# To deflate historical USD values, FAO uses the WB MUV series.
# Let's do the same.
# Available on the "Pink Sheet" (annual) https://www.worldbank.org/en/research/commodity-markets
thisFile <- "CMO-Historical-Data-Annual.xlsx"
thisFilepath <- paste0(thisFolder, thisFile)
dfMUV <- readxl::read_xlsx(thisFilepath, sheet = "Annual Indices (Real)")
dfMUV <- dfMUV %>% select(year = 1, MUV = 18) %>% slice(-c(1:8))
dfMUV$MUV <- as.numeric(dfMUV$MUV)
# Rebase MUV to more suitable year dollars (default is 2010 dollars)
# (Rebase to match FAO 2014-2016 constant USD used in FAO VAP data)
muvRebaser <- dfMUV %>%
  filter(year %in% c("2014", "2015", "2016")) %>%
  summarise(muvRebaser = mean(MUV)) %>%
  pull(muvRebaser)
df <- df %>% merge(dfMUV) %>% mutate(priceSeries = price * muvRebaser / MUV)
# # Get historical CPI deflator series (Historical CPI-U) from Bureao of Labor Statistics
# #https://www.bls.gov/cpi/tables/supplemental-files/
# # Use "Index Averages" worksheet (the annual avg over months)
# # Note this deflates to 1982-84 dollars. To rebase to 2024 dollars: CPIrebased = cpiSeries / (cpi2024value / 100)
# thisFile <- "historical-cpi-u-202512.xlsx"
# thisFilepath <- paste0(thisFolder, thisFile)
# dfCPI <- readxl::read_excel(thisFilepath, sheet = "Index Averages")
# dfCPI <- dfCPI %>% select(year = 2, avgCPI = 5) %>% slice(-c(1:5))
# dfCPI$avgCPI <- as.numeric(dfCPI$avgCPI)
# # Rebase CPI to more suitable year dollars (default is 1982-84 dollars)
# # (Rebase to match FAO 2014-2016 constant USD used in VAP data)
# cpiRebaser <- dfCPI %>%
#   filter(year %in% c("2014", "2015", "2016")) %>%
#   summarise(cpiRebaser = mean(avgCPI)) %>%
#   pull(cpiRebaser)
# df <- df %>% merge(dfCPI) %>% mutate(priceSeries = price * cpiRebaser / avgCPI)
#----------------------------------------------------------------------------
# Check stationarity
# library(Kendall)
# df <- df %>% group_by(Item) %>% mutate(MK_kgPrice = as.numeric(MannKendall(priceSeries)[1]))
# hist(df$MK_kgPrice)
# library(fable)
# library(feasts)
# library(tsibble)
# library(features)
# # 1. Convert your df to a tsibble
# df_ts <- df %>%
#   mutate(year = as.integer(year)) %>%
#   as_tsibble(index = year, key = Item)
# # 2. Calculate features
# df_features <- df_ts %>% features(priceSeries, feat_stl)
# # 3. Look at the 'trend_strength' column
# # This will give you a value 0-1 for every commodity
# head(df_features)
#----------------------------------------------------------------------------
# Plot a random sample of deflated commodity world export price series
# num_to_plot <- 5
# randomItems <- sample(unique(df$Item), num_to_plot)
#df$Item[grep("Rice", df$Item)] %>% unique()
plotThese <- c("Wheat", "Rice, milled", "Beans, dry", "Natural Rubber", "Beef and Buffalo Meat, primary", "Sugar Crops Primary")
yTitle <- "World export real price\n(2014-2016 USD / kg)"
plotTitle <- "World export real price series\n(1961-2024) for selected commodities"
dfPlotTs <- df %>% mutate(year = as.integer(year)) %>% subset(Item %in% plotThese)
gg <- plotPriceSeries(thisDfPlotTs = dfPlotTs, plotTitle, yTitle)
ggSampleTsExpPrice <- gg
#----------------------------------------------------------------------------
# Risk reward plot
dfRiskReward <- processRiskRewardDf(df)
dfRiskReward <- dfRiskReward %>% subset(group != "Other" & Item != "Crude Materials nes")
# Set group colors to be used in all risk reward plots
n <- length(unique(dfRiskReward$group))
bag_of_colors <- randomcoloR::distinctColorPalette(k = 2 * n)
colorVec <- sample(bag_of_colors, n)
#----------------------------------------------------------------------------
# Get slope and intercept of the risk reward relationship
outVec <- getSlopeEtc(dfRiskReward)
m <- outVec[1]; b <- outVec[2]; adjR2 <- outVec[3]; N <- outVec[4];
#----------------------------------------------------------------------------
#dfPlot$Item[grep("Maize", dfPlot$Item, ignore.case = T)]
labelItems <- c("Maize (corn)", "Flour of maize",
  #"Wheat", "Flour, wheat",
  #"Starch, cassava",
                "Coffee, green", "Coffee, decaffeinated or roasted", #"Bananas",
                #"Plantains",
                #"Sweet potatoes",
                #"Potatoes",
                "Cocoa beans",
                #"Wool, greasy", 
                #"Beans, dry",
                "Raw silk (not thrown)",
                #"Apples",
  "Vanilla, raw",
                "Soya beans", "Soya bean oil", "Natural rubber", "Rice",
                "Cocoa butter, fat and oil")
#---
thisTitle <- "World export real price"
thisXlab <- NULL; thisYlab <- "Ln(mean)"
ggRiskRewardExpPrice <- plotRiskReward(dfRiskReward, labelItems, thisTitle, thisXlab, thisYlab, legendOn = F, colorVec, m, b, N, adjR2)
#============================================================================
# Farmgate price series
#============================================================================
thisFile <- "Production_Crops_Livestock_E_All_Data.csv"
thisFilepath <- paste0(thisFolder, thisFile)
theseElements <- c("Production", "Area harvested", "Yield"); theseUnits <- c("t", "ha", "kg/ha")
dfAreaProd <- processDf(thisFilepath, theseElements, theseUnits, isExpPrice = F) %>%
  spread(Element, val) %>% select(Item, year, Production, `Area harvested`, Yield)
thisFile <- "Value_of_Production_E_All_Data.csv"
thisFilepath <- paste0(thisFolder, thisFile)
theseElements <- "Gross Production Value (constant 2014-2016 thousand US$)"; theseUnits <- "1000 USD"
dfVAP <- processDf(thisFilepath, theseElements, theseUnits, isExpPrice = F) %>%
  select(Item, year, val) %>% rename(VAP = val); dfVAP$Element <- NULL
df <- dfAreaProd %>% merge(dfVAP, by = c("Item", "year")) %>%
  mutate(haPrice = 1000 * VAP / `Area harvested`, kgPrice = VAP / Production,
         year = as.integer(year))
#----------------------------------------------------------------------------
# Check stationarity
# library(Kendall)
# # Example for one item:
# df <- df %>% group_by(Item) %>% mutate(MK_kgPrice = as.numeric(MannKendall(kgPrice)[1]), MK_haPrice = as.numeric(MannKendall(haPrice)[1]))
# hist(df$MK_kgPrice)
# hist(df$MK_haPrice)
# x <- MannKendall(df$kgPrice[df$Item == "Abaca, manila hemp, raw"])
#----------------------------------------------------------------------------
# Plot a random sample of deflated commodity world price series
#df$Item[grep("Soya beans", df$Item, ignore.case = T)] %>% unique()
#"Coffee, green", "Wheat", "Soya beans",
plotThese <- c("Sugar Crops Primary", "Walnuts, in shell",
               "Pistachios, in shell", "Beeswax",
              "Raspberries", "Sesame seed")
# num_to_plot <- 15
# randomItems <- sample(unique(df$Item), num_to_plot)
# plotThese <- randomItems
yTitle <-"World producer real price\n(2014-2016 USD / kg)"
plotTitle <- "World producer real price series\n(1961-2024) for selected commodities (constant 2014-2016 USD)"
dfPlotTs <- df %>% subset(Item %in% plotThese) %>% rename(priceSeries = kgPrice)
gg <- plotPriceSeries(dfPlotTs, plotTitle, yTitle)
ggSampleTsProdPrice <- gg
yTitle <-"World producer real price per hectare\n(2014-2016 USD / hectare)"
plotTitle <- "World producer real production value per hectare series\n(1961-2024) for selected commodities (constant 2014-2016 USD)"
dfPlotTs <- df %>% subset(Item %in% plotThese) %>% rename(priceSeries = haPrice)
gg <- plotPriceSeries(dfPlotTs, plotTitle, yTitle)
ggSampleTsProdHaPrice <- gg
#----------------------------------------------------------------------------
# Risk reward plot
dfKg <- df %>% rename(priceSeries = kgPrice)
dfRiskReward <- processRiskRewardDf(dfKg)
dfRiskReward <- dfRiskReward %>% subset(group != "Other")
#----------------------------------------------------------------------------
# Get slope and intercept of the risk reward relationship
outVec <- getSlopeEtc(dfRiskReward)
m <- outVec[1]; b <- outVec[2]; adjR2 <- outVec[3]; N <- outVec[4];
#----------------------------------------------------------------------------
labelItems <- c("Maize (corn)",
                "Coffee, green",
                "Bananas",
                #"Plantains",
                #"Sweet potatoes",
                #"Potatoes",
                "Cocoa beans",
                #"Wool, greasy", 
                "Beans, dry",
                "Silk-worm cocoons suitable for reeling",
                #"Apples",
                "Vanilla, raw",
                "Soya beans", "Natural rubber in primary forms", "Rice")
thisTitle <- "World producer real price/kg "
thisYlab <- NULL; thisXlab <- NULL
ggRiskRewardProdKgPrice <- plotRiskReward(dfRiskReward, labelItems, thisTitle, thisXlab, thisYlab, legendOn = F, colorVec, m, b, N, adjR2)
#----------------------------------------------------------------------------
dfYd <- df %>% rename(priceSeries = Yield)
dfRiskReward <- processRiskRewardDf(dfYd)
dfRiskReward <- dfRiskReward %>% subset(group != "Other")
#----------------------------------------------------------------------------
# Get slope and intercept of the risk reward relationship
outVec <- getSlopeEtc(dfRiskReward)
m <- outVec[1]; b <- outVec[2]; adjR2 <- outVec[3]; N <- outVec[4];
#----------------------------------------------------------------------------
thisTitle <- "World yield"
thisXlab <- "Ln(stand. dev.)"; thisYlab <- NULL
ggRiskRewardYd <- plotRiskReward(dfRiskReward, labelItems, thisTitle, thisXlab, thisYlab, legendOn = F, colorVec, m, b, N, adjR2)
#----------------------------------------------------------------------------
dfHa <- df %>% rename(priceSeries = haPrice)
dfRiskReward <- processRiskRewardDf(dfHa)
dfRiskReward <- dfRiskReward %>% subset(group != "Other")
#----------------------------------------------------------------------------
# Get slope and intercept of the risk reward relationship
outVec <- getSlopeEtc(dfRiskReward)
m <- outVec[1]; b <- outVec[2]; adjR2 <- outVec[3]; N <- outVec[4];
#----------------------------------------------------------------------------
thisTitle <- "World producer real price/ha"
thisYlab <- "Ln(mean)"; thisXlab <- "Ln(stand. dev.)"
ggRiskRewardProdHaPrice <- plotRiskReward(dfRiskReward, labelItems, thisTitle, thisXlab, thisYlab, legendOn = T, colorVec, m, b, N, adjR2)
#===============================================================================
# Put together the 2x2 display of panels
# 1. Arrange the plots in a 2x2 grid
# The "/" operator stacks plots vertically, while "|" or "+" aligns them horizontally
combined_plot <- (ggRiskRewardExpPrice | ggRiskRewardProdKgPrice) / (ggRiskRewardProdHaPrice | ggRiskRewardYd)

# 2. Add labels and a global theme
(combined_plot + 
  plot_annotation(
    tag_levels = 'A',                      # Use uppercase letters A, B, C, D
    title = "Empirical evidence of a reward-stability tradeoff in commodity prices and yields",
    subtitle = "FAO price and yield series 1961-2024, dollar values in constant 2014-2016 USD",
    #caption = "Source: Pathogen Model v1.2"
  )) & 
  theme(plot.tag = element_text(face = 'bold')) # Apply bold to all tags at once

```

# Empirical evidence of a reward-stability tradeoff in agricultural commodities

To assess the relationship between agricultural risk and reward, the mean and standard deviation of global yield and price series over 1961-2024 are plotted against each other in logs (Figure \ref{fig:riskReward}). Only commodities with a complete, non-zero time series are included. Comparisons of moments across time series are meaningful insofar as the series are stationary. To ensure stationarity, standard deviations are computed from first-differenced series. In the price series (panels A-C), nominal values are deflated using the World Bank Manufacturers Unit Value Index [@WB_MUV_2026], such that the means are expressed in constant US dollars. The means of yield series (panel D), however, should be interpreted with great caution, since yield series generally exhibit a secular trend. Producer price is displayed in both per unit weight and per unit area harvested terms, indicating that the observed regularity is robust to alternative value denominations.

The plots reveal a strong empirical correspondence between level and volatility across a wide range of commodity types, including crop, animal, food, non-food, raw, and processed commodities. The formal economic mechanism behind the regularity, if any exists, does not readily suggest itself. Similar scaling relations have often been observed in other contexts---primarily biological and ecological systems---where they are known as Taylor’s Law or fluctuation scaling [@eisler2008fluctuation]. However, the explanations offered in those contexts are domain specific and do not readily extend to agricultural commodity markets. The search for abstract statistical mechanisms not rooted in any particular empirical context is an active area of research [@cohen2015random; @brown2021taylor; @taylor2019taylor].
<!-- Notably, the scaling relation appears consistently across commodities with markedly different biological, technological, and market characteristics. -->

# Theoretical basis for a yield level-stability tradeoff

In the cultivation of any single agricultural commodity, it is straightforward to demonstrate a fundamental theoretical reason to expect a tradeoff between yield level and yield stability.

Under the standard assumption that yield $y(x)$ is a homogeneous function of inputs $x$ of degree $h$, Euler’s theorem implies

\begin{equation}
\sum_{i=1}^n x_i \frac{\partial y}{\partial x_i}
= h y(\mathbf{x})
\end{equation}

or equivalently,

\begin{equation}
\mathbf{x}' \nabla y(\mathbf{x})
= h y(\mathbf{x}).
\end{equation}

Differentiating both sides yields

\begin{equation}
\nabla \left( \mathbf{x}' \nabla y \right)
= \nabla (h y).
\end{equation}

Applying the product rule,

\begin{equation}
\nabla y + H(\mathbf{x}) \mathbf{x}
= h \nabla y,
\end{equation}

where $H(x)$ denotes the Hessian matrix.

Rearranging,

\begin{equation}
H(\mathbf{x}) \mathbf{x}
= (h - 1)\nabla y.
\end{equation}

Premultiplying by $x'$

\begin{equation}
\mathbf{x}' H(\mathbf{x}) \mathbf{x}
= (h - 1) \mathbf{x}' \nabla y.
\end{equation}

Invoking Euler’s theorem once more,

\begin{equation}
\mathbf{x}' H(\mathbf{x}) \mathbf{x}
= h(h - 1) y(\mathbf{x}).
\label{eq:volatPropToLev}
\end{equation}

The quadratic form $x'H(x) x$ captures local curvature of the yield function and therefore measures sensitivity of output to small perturbations of inputs.

For a strictly concave production technology, the Hessian is negative definite, implying

\begin{equation}
\mathbf{x}' H(\mathbf{x}) \mathbf{x} < 0.
\end{equation}

This condition is satisfied when $0 < h < 1$, corresponding to decreasing returns to scale. Equation \ref{eq:volatPropToLev} therefore implies that curvature---and hence local input sensitivity---scales proportionately with yield level.

In economic terms, higher-yield production regimes are inherently more sensitive to deviations of inputs from their optimal values, providing a theoretical basis for a yield level–stability tradeoff.

# Conclusion

The empirical regularity and theoretical argument presented here suggest that agricultural yield and price volatility cannot be attributed solely to exogenous random shocks. Instead, variability appears to exhibit systematic scaling behavior, arising from scale-dependent sensitivity to exogenous disturbances.
<!-- This mechanism provides a parsimonious interpretation of the observed reward–stability tradeoff. -->

Note that the empirical results are derived from world-level aggregates. Future work may examine the extent to which these regularities persist under geographical disaggregation, where deviations or slope heterogeneity may offer additional insights.

From an AR4D policy perspective, the results imply that the pursuit of higher yields and greater participation in high-value commodity markets may conflict with objectives centered on yield stability and smallholder resilience. Yield enhancement strategies may therefore benefit from explicit consideration of risk-bearing capacity within target populations and production environments.

In this context, yield-enhancing technologies and practices may be most effective when complemented by mechanisms that mitigate exposure to variability, including financial instruments, institutional arrangements, and cooperative risk-pooling structures. Frameworks emphasizing joint productivity and stability objectives may thus benefit from a more explicit recognition of scale-dependent variability.

# References

<!-- The empirical regularity and theoretical argument outlined above indicate that agricultural yield and price volatility are not due to random exogenous shocks alone, but rather arise in a partly deterministic/endogenous/mechanical [which word to use?] manner from scale-dependent sensitivity to random shocks---resulting in the observed reward-stability tradeoff. It is important to keep in mind that the empirical findings are world level. A more refined analysis can examine to what degree the regularity persists or whether anything interesting happens (e.g. changing slope) under geographical disaggregation. -->
<!-- For AR4D policy purposes, this means that the pursuit of increased yields and smallholder exposure to high value commodity markets may work against the simultaneous pursuit of higher yield stability and overall smallholder resilience. AR4D yield enhancement strategy should thus start from a careful assessment of the risk capital available in target populations and environments; and complement yield enhancing technology and practices accordingly with the development of appropriate risk mitigating mechanisms (e.g. microfinance, subsidies, insurance, cooperative social mechanisms). Modern AR4D frameworks emphasizing both higher yields and higher stability such as the "Agricultural Innovation Systems" framework [@sandlin2014agricultural; @klerkx2010adaptive] or the "Blended Sustainability" vision [@mockshell2018beyond] can thus take a more grounded, informed approach. -->
<!-- Above I document an empirical tradeoff between stability and reward in world commodity markets, and demonstrate theoretically that, at farm level, yield volatility scales with yield level under standard microeconomic assumptions regarding homogeneity and concavity. This strongly -->
<!-- In order to assist authors in the process of preparing a manuscript for a journal, the Taylor & Francis '\textsf{Interact}' layout style has been implemented as a \LaTeXe\ class file based on the `article` document class. A \textsc{Bib}\TeX\ bibliography style file and a sample bibliography are also provided in order to assist with the formatting of your references. -->

<!-- Commands that differ from or are provided in addition to standard \LaTeXe\ are described in this document, which is _not_ a substitute for a \LaTeXe\ tutorial. -->

<!-- The `interactcadsample.tex` file can be used as a template for a manuscript by cutting, pasting, inserting and deleting text as appropriate, using the preamble and the \LaTeX\ environments provided (e.g.\ `\begin{abstract}`, `\begin{keywords}`). -->

<!-- ## The \textsf{Interact} class file -->

<!-- The `interact` class file preserves the standard \LaTeXe\ interface such that any document that can be produced using `article.cls` can also be produced with minimal alteration using the `interact` class file as described in this document. -->

<!-- If your article is accepted for publication it will be typeset as the journal requires in Minion Pro and/or Myriad Pro. Since most authors will not have these fonts installed, the page make-up is liable to alter slightly with the change of font. Also, the `interact` class file produces only single-column format, which is preferred for peer review and will be converted to two-column format by the typesetter if necessary during preparation of the proofs. Please therefore do not try to match the typeset format exactly, but use the standard \LaTeX\ fonts instead and ignore details such as slightly long lines of text or figures/tables not appearing in exact synchronization with their citations in the text: these details will be dealt with by the typesetter. Similarly, it is unnecessary to spend time addressing warnings in the log file -- if your .tex file compiles to produce a PDF document that correctly shows how you wish your paper to appear, such warnings will not prevent your source files being imported into the typesetter's program. -->

<!-- ## Submission of manuscripts prepared using _\LaTeX_ -->

<!-- Manuscripts for possible publication should be submitted to the Editors for review as directed in the journal's Instructions for Authors, and in accordance with any technical instructions provided in the journal's ScholarOne Manuscripts or Editorial Manager site. Your \LaTeX\ source file(s), the class file and any graphics files will be required in addition to the final PDF version when final, revised versions of accepted manuscripts are submitted. -->

<!-- Please ensure that any author-defined macros used in your article are gathered together in the preamble of your .tex file, i.e.\ before the `\begin{document}` command. Note that if serious problems are encountered in the coding of a document (missing author-defined macros, for example), the typesetter may resort to rekeying it. -->

<!-- # Using the `interact` class file -->

<!-- For convenience, simply copy the `interact.cls` file into the same directory as your manuscript files (you do not need to install it in your \TeX\ distribution). In order to use the `interact` document class, replace the command `\documentclass{article}` at the beginning of your document with the command `\documentclass{interact}`. -->

<!-- The following document-class options should _not_ be used with the `interact` class file: -->

<!-- * `10pt`, `11pt`, `12pt` -- unavailable; -->
<!-- * `oneside`, `twoside` -- not necessary, `oneside` is the default; -->
<!-- * `leqno`, `titlepage` -- should not be used; -->
<!-- * `twocolumn` -- should not be used; -->
<!-- * `onecolumn` -- not necessary as it is the default style. -->

<!-- To prepare a manuscript for a journal that is printed in A4 (two column) format, use the `largeformat` document-class option provided by `interact.cls`; otherwise the class file produces pages sized for B5 (single column) format by default. The `geometry` package should not be used to make any further adjustments to the page dimensions. -->

<!-- <!-- If your manuscript has supplementary content you can prepare this using the `suppldata` document-class option, which will suppress the `article history' date. This option must _not_ be used on any primary content. --> -->

<!-- # Additional features of the `interact` class file -->

<!-- ## Title, authors' names and affiliations, abstracts and article types -->

<!-- ## Abbreviations -->

<!-- A list of abbreviations may be included if required, enclosed within an `abbreviations` environment, i.e.\ `\begin{abbreviations}`\ldots`\end{abbreviations}`, immediately following the `abstract` environment. -->


<!-- ## Keywords -->

<!-- A list of keywords may be included if required, enclosed within a `keywords` environment, i.e.\ `\begin{keywords}`\ldots`\end{keywords}`. Additional keywords in other languages (preceded by a translation of the word `keywords`) may also be included within the `keywords` environment if required. -->


<!-- ## Subject classification codes -->

<!-- AMS, JEL or PACS classification codes may be included if required. The `interact` class file provides an `amscode` environment, i.e.\ `\begin{amscode}`\ldots`\end{amscode}`, a `jelcode` environment, i.e.\ `\begin{jelcode}`\ldots`\end{jelcode}`, and a `pacscode` environment, i.e.\ `\begin{pacscode}`\ldots`\end{pacscode}` to assist with this. -->


<!-- ## Additional footnotes to the title or authors' names -->

<!-- <!-- The `\thanks` command may be used to create additional footnotes to the title or authors' names if required. Footnote symbols for this purpose should be used in the order --> -->
<!-- <!-- $^\ast$~(coded as `$^\ast$`), $\dagger$~(`$\dagger$`), $\ddagger$~(`$\ddagger$`), $\S$~(`$\S$`), $\P$~(`$\P$`), $\|$~(`$\|$`), --> -->
<!-- <!-- $\dagger\dagger$~(`$\dagger\dagger$`), $\ddagger\ddagger$~(`$\ddagger\ddagger$`), $\S\S$~(`$\S\S$`), $\P\P$~(`$\P\P$`). --> -->

<!-- Note that any `footnote`s to the main text will automatically be assigned the superscript symbols 1, 2, 3, etc. by the class file.^[If preferred, the `endnotes` package may be used to set the notes at the end of your text, before the bibliography. The symbols will be changed to match the style of the journal if necessary by the typesetter.] -->

<!-- # Embedding R code -->

<!-- ## Code chunks -->


<!-- ```{r cars} -->
<!-- summary(cars) -->
<!-- ``` -->

<!-- ## Including Plots -->

<!-- You can also embed plots, for example: -->

<!-- ```{r pressure, echo=FALSE, out.width='80%'} -->
<!-- plot(pressure) -->
<!-- ``` -->

<!-- Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot. -->

<!-- # Some guidelines for using the standard features of \LaTeX -->

<!-- ## Sections -->

<!-- The \textsf{Interact} layout style allows for five levels of section heading, all of which are provided in the `interact` class file using the standard \LaTeX\ commands `\section`, `\subsection`, `\subsubsection`, `\paragraph` and `\subparagraph`. Numbering will be automatically generated for all these headings by default. -->

<!-- ## Lists -->

<!-- Numbered lists are produced using the `enumerate` environment, which will number each list item with arabic numerals by default. For example, -->

<!-- 1. first item -->
<!-- 2. second item -->
<!-- 3. third item -->

<!-- Alternative numbering styles can be achieved by inserting an optional argument in square brackets to each `item`, e.g.\ `\item[(i)] first item`\, to create a list numbered with roman numerals at level one. -->

<!-- Bulleted lists are produced using the `itemize` environment. For example, -->

<!-- * First bulleted item -->
<!-- * Second bulleted item -->
<!-- * Third bulleted item -->

<!-- ## Figures -->

<!-- ```{r pressure-plot, fig.show='hide'} -->
<!-- plot(pressure) -->
<!-- ``` -->

<!-- The `interact` class file will deal with positioning your figures in the same way as standard \LaTeX. It should not normally be necessary to use the optional `[htb]` location specifiers of the `figure` environment in your manuscript; you may, however, find the `[p]` placement option or the `endfloat` package useful if a journal insists on the need to separate figures from the text. -->

<!-- Figure captions appear below the figures themselves, therefore the `\caption` command should appear after the body of the figure. For example, Figure~\ref{sample-figure} with caption and sub-captions is produced using the following commands: -->
<!-- ``` -->
<!-- \begin{figure} -->
<!-- \centering -->
<!-- \subfigure[An example of an individual figure sub-caption.]{% -->
<!-- \resizebox*{5cm}{!}{\includegraphics{path/to/fig}}}\hspace{5pt} -->
<!-- \subfigure[A slightly shorter sub-caption.]{% -->
<!-- \resizebox*{5cm}{!}{\includegraphics{path/to/fig}}} -->
<!-- \caption{Example of a two-part figure with individual sub-captions -->
<!--  showing that captions are flush left and justified if greater -->
<!--  than one line of text.} \label{sample-figure} -->
<!-- \end{figure} -->
<!-- ``` -->
<!-- \begin{figure} -->
<!-- \centering -->
<!-- \subfigure[An example of an individual figure sub-caption.]{% -->
<!-- \resizebox*{5cm}{!}{\includegraphics{`r knitr::fig_chunk('pressure-plot', 'pdf')`}}}\hspace{5pt} -->
<!-- \subfigure[A slightly shorter sub-caption.]{% -->
<!-- \resizebox*{5cm}{!}{\includegraphics{`r knitr::fig_chunk('pressure-plot', 'pdf')`}}} -->
<!-- \caption{Example of a two-part figure with individual sub-captions -->
<!--  showing that captions are flush left and justified if greater -->
<!--  than one line of text.} \label{sample-figure} -->
<!-- \end{figure} -->

<!-- To ensure that figures are correctly numbered automatically, the `\label` command should be included just after the `\caption` command, or in its argument. -->

<!-- The `\subfigure` command requires `subfigure.sty`, which is called in the preamble of the `interacttfssample.tex` file (to allow your choice of an alternative package if preferred) and included in the \textsf{Interact} \LaTeX\ bundle for convenience. Please supply any additional figure macros used with your article in the preamble of your .tex file. -->

<!-- The source files of any figures will be required when the final, revised version of a manuscript is submitted. Authors should ensure that these are suitable (in terms of lettering size, etc.) for the reductions they envisage. -->

<!-- The `epstopdf` package can be used to incorporate encapsulated PostScript (.eps) illustrations when using PDF\LaTeX, etc. Please provide the original .eps source files rather than the generated PDF images of those illustrations for production purposes. -->


<!-- ## Tables -->

<!-- The `interact` class file will deal with positioning your tables in the same way as standard \LaTeX. It should not normally be necessary to use the optional `[htb]` location specifiers of the `table` environment in your manuscript; you may, however, find the `[p]` placement option or the `endfloat` package useful if a journal insists on the need to separate tables from the text. -->

<!-- The `tabular` environment can be used as shown to create tables with single horizontal rules at the head, foot and elsewhere as appropriate. The captions appear above the tables in the \textsf{Interact} style, therefore the `\tbl` command should be used before the body of the table. For example, Table~\ref{sample-table} is produced using the following commands: -->
<!-- \begin{table} -->
<!-- \tbl{Example of a table showing that its caption is as wide as -->
<!--  the table itself and justified.} -->
<!-- {\begin{tabular}{lcccccc} \toprule -->
<!--  & \multicolumn{2}{l}{Type} \\ \cmidrule{2-7} -->
<!--  Class & One & Two & Three & Four & Five & Six \\ \midrule -->
<!--  Alpha\textsuperscript{a} & A1 & A2 & A3 & A4 & A5 & A6 \\ -->
<!--  Beta & B2 & B2 & B3 & B4 & B5 & B6 \\ -->
<!--  Gamma & C2 & C2 & C3 & C4 & C5 & C6 \\ \bottomrule -->
<!-- \end{tabular}} -->
<!-- \tabnote{\textsuperscript{a}This footnote shows how to include -->
<!--  footnotes to a table if required.} -->
<!-- \label{sample-table} -->
<!-- \end{table} -->
<!-- ``` -->
<!-- \begin{table} -->
<!-- \tbl{Example of a table showing that its caption is as wide as -->
<!--  the table itself and justified.} -->
<!-- {\begin{tabular}{lcccccc} \toprule -->
<!--  & \multicolumn{2}{l}{Type} \\ \cmidrule{2-7} -->
<!--  Class & One & Two & Three & Four & Five & Six \\ \midrule -->
<!--  Alpha\textsuperscript{a} & A1 & A2 & A3 & A4 & A5 & A6 \\ -->
<!--  Beta & B2 & B2 & B3 & B4 & B5 & B6 \\ -->
<!--  Gamma & C2 & C2 & C3 & C4 & C5 & C6 \\ \bottomrule -->
<!-- \end{tabular}} -->
<!-- \tabnote{\textsuperscript{a}This footnote shows how to include -->
<!--  footnotes to a table if required.} -->
<!-- \label{sample-table} -->
<!-- \end{table} -->
<!-- ``` -->

<!-- To ensure that tables are correctly numbered automatically, the `\label` command should be included just before `\end{table}`. -->

<!-- The `\toprule`, `\midrule`, `\bottomrule` and `\cmidrule` commands are those used by `booktabs.sty`, which is called by the `interact` class file and included in the \textsf{Interact} \LaTeX\ bundle for convenience. Tables produced using the standard commands of the `tabular` environment are also compatible with the `interact` class file. -->


<!-- ## Landscape pages -->

<!-- If a figure or table is too wide to fit the page it will need to be rotated, along with its caption, through 90$^{\circ}$ anticlockwise. Landscape figures and tables can be produced using the `rotating` package, which is called by the `interact` class file. The following commands (for example) can be used to produce such pages. -->
<!-- ``` -->
<!-- \setcounter{figure}{1} -->
<!-- \begin{sidewaysfigure} -->
<!-- \centerline{\epsfbox{figname.eps}} -->
<!-- \caption{Example landscape figure caption.} -->
<!-- \label{landfig} -->
<!-- \end{sidewaysfigure} -->
<!-- ``` -->
<!-- ``` -->
<!-- \setcounter{table}{1} -->
<!-- \begin{sidewaystable} -->
<!--  \tbl{Example landscape table caption.} -->
<!--   {\begin{tabular}{@{}llllcll} -->
<!--     . -->
<!--     . -->
<!--     . -->
<!--   \end{tabular}}\label{landtab} -->
<!-- \end{sidewaystable} -->
<!-- ``` -->
<!-- Before any such float environment, use the `\setcounter` command as above to fix the numbering of the caption (the value of the counter being the number given to the preceding figure or table). Subsequent captions will then be automatically renumbered accordingly. The `\epsfbox` command requires `epsfig.sty`, which is called by the `interact` class file and is also included in the \textsf{Interact} \LaTeX\ bundle for convenience. -->

<!-- Please note that if the `endfloat` package is used, one or both of the commands -->
<!-- ``` -->
<!-- \DeclareDelayedFloatFlavor{sidewaysfigure}{figure} -->
<!-- \DeclareDelayedFloatFlavor{sidewaystable}{table} -->
<!-- ``` -->
<!-- will need to be included in the preamble of your .tex file, after the `endfloat` package is loaded, in order to process any landscape figures and/or tables correctly. -->


<!-- ## Theorem-like structures -->

<!-- A predefined `proof` environment is provided by the `amsthm` package (which is called by the `interact` class file), as follows: -->
<!-- \begin{proof} -->
<!-- More recent algorithms for solving the semidefinite programming relaxation are particularly efficient, because they explore the structure of the MAX-CUT problem. -->
<!-- \end{proof} -->
<!-- \noindent This was produced by simply typing: -->
<!-- ``` -->
<!-- \begin{proof} -->
<!-- More recent algorithms for solving the semidefinite programming -->
<!-- relaxation are particularly efficient, because they explore the -->
<!-- structure of the MAX-CUT problem. -->
<!-- \end{proof} -->
<!-- ``` -->
<!-- Other theorem-like environments (theorem, definition, remark, etc.) need to be defined as required, e.g.\ using `\newtheorem{theorem}{Theorem}` in the preamble of your .tex file (see the preamble of `interactcadsample.tex` for more examples). You can define the numbering scheme for these structures however suits your article best. Please note that the format of the text in these environments may be changed if necessary to match the style of individual journals by the typesetter during preparation of the proofs. -->


<!-- ## Mathematics -->

<!-- ### Displayed mathematics -->

<!-- The `interact` class file will set displayed mathematical formulas centred on the page without equation numbers if you use the `displaymath` environment or the equivalent `\[...\]` construction. For example, the equation -->
<!-- \[ -->
<!--  \hat{\theta}_{w_i} = \hat{\theta}(s(t,\mathcal{U}_{w_i})) -->
<!-- \] -->
<!-- was typeset using the commands -->
<!-- ``` -->
<!-- \[ -->
<!--  \hat{\theta}_{w_i} = \hat{\theta}(s(t,\mathcal{U}_{w_i})) -->
<!-- \] -->
<!-- ``` -->

<!-- For those of your equations that you wish to be automatically numbered sequentially throughout the text for future reference, use the `equation` environment, e.g. -->
<!-- \begin{equation} -->
<!--  \hat{\theta}_{w_i} = \hat{\theta}(s(t,\mathcal{U}_{w_i})) -->
<!-- \end{equation} -->
<!-- was typeset using the commands -->
<!-- ``` -->
<!-- \begin{equation} -->
<!--  \hat{\theta}_{w_i} = \hat{\theta}(s(t,\mathcal{U}_{w_i})) -->
<!-- \end{equation} -->
<!-- ``` -->

<!-- Part numbers for sets of equations may be generated using the `subequations` environment, e.g. -->
<!-- \begin{subequations} \label{subeqnexample} -->
<!-- \begin{equation} -->
<!--      \varepsilon \rho w_{tt}(s,t) = N[w_{s}(s,t),w_{st}(s,t)]_{s}, -->
<!--      \label{subeqnparta} -->
<!-- \end{equation} -->
<!-- \begin{equation} -->
<!--      w_{tt}(1,t)+N[w_{s}(1,t),w_{st}(1,t)] = 0, -->
<!--      \label{subeqnpartb} -->
<!-- \end{equation} -->
<!-- \end{subequations} -->
<!-- which was typeset using the commands -->
<!-- ``` -->
<!-- \begin{subequations} \label{subeqnexample} -->
<!-- \begin{equation} -->
<!--      \varepsilon \rho w_{tt}(s,t) = N[w_{s}(s,t),w_{st}(s,t)]_{s}, -->
<!--      \label{subeqnparta} -->
<!-- \end{equation} -->
<!-- \begin{equation} -->
<!--      w_{tt}(1,t)+N[w_{s}(1,t),w_{st}(1,t)] = 0,   \label{subeqnpartb} -->
<!-- \end{equation} -->
<!-- \end{subequations} -->
<!-- ``` -->
<!-- <!-- This is made possible by the `amsmath` package, which is called by the class file. If you put a `\label` just after the `\begin{subequations}` command, references can be made to the collection of equations, i.e.\ `(\ref{subeqnexample})' in the example above. Or, as the example also shows, you can label and refer to each equation individually -- i.e.\ `(\ref{subeqnparta})' and `(\ref{subeqnpartb})'. --> -->

<!-- Displayed mathematics should be given end-of-line punctuation appropriate to the running text sentence of which it forms a part, if required. -->

<!-- ### Math fonts -->

<!-- #### Superscripts and subscripts -->
<!-- Superscripts and subscripts will automatically come out in the correct size in a math environment (i.e.\ enclosed within `\(...\)` or `$...$` commands in running text, or within `\[...\]` or the `equation` environment for displayed equations). Sub/superscripts that are physical variables should be italic, whereas those that are labels should be roman (e.g.\ $C_p$, $T_\mathrm{eff}$). If the subscripts or superscripts need to be other than italic, they must be coded individually. -->

<!-- #### Upright Greek characters and the upright partial derivative sign -->
<!-- Upright lowercase Greek characters can be obtained by inserting the letter `u` in the control code for the character, e.g.\ `\umu` and `\upi` produce $\umu$ (used, for example, in the symbol for the unit microns -- $\umu\mathrm{m}$) and $\upi$ (the ratio of the circumference of a circle to its diameter). Similarly, the control code for the upright partial derivative $\upartial$ is `\upartial`. Bold lowercase as well as uppercase Greek characters can be obtained by `{\bm \gamma}`, for example, which gives ${\bm \gamma}$, and `{\bm \Gamma}`, which gives ${\bm \Gamma}$. -->


<!-- # Acknowledgement(s) {-} -->

<!-- An unnumbered section, e.g.\ `\section*{Acknowledgements}`, may be used for thanks, etc.\ if required and included _in the non-anonymous version_ before any Notes or References. -->

<!-- # Disclosure statement {-} -->

<!-- An unnumbered section, e.g.\ `\section*{Disclosure statement}`, may be used to declare any potential conflict of interest and included _in the non-anonymous version_ before any Notes or References, after any Acknowledgements and before any Funding information. -->

<!-- # Funding {-} -->

<!-- An unnumbered section, e.g.\ `\section*{Funding}`, may be used for grant details, etc.\ if required and included _in the non-anonymous version_ before any Notes or References. -->

<!-- # Notes on contributor(s) {-} -->

<!-- An unnumbered section, e.g.\ `\section*{Notes on contributors}`, may be included _in the non-anonymous version_ if required. A photograph may be added if requested. -->

<!-- # Nomenclature/Notation {-} -->

<!-- An unnumbered section, e.g.\ `\section*{Nomenclature}` (or `\section*{Notation}`), may be included if required, before any Notes or References. -->

<!-- # Notes {-} -->

<!-- An unnumbered `Notes` section may be included before the References (if using the `endnotes` package, use the command `\theendnotes` where the notes are to appear, instead of creating a `\section*`). -->

<!-- # References -->

<!-- ## References cited in the text -->

<!-- ## The list of references -->

<!-- References should be listed at the end of the main text in alphabetical order by authors' surnames, then chronologically (earliest first). If references have the same author(s), editor(s), etc., arrange by year of publication, with undated works at the end. -->
<!-- A single-author entry precedes a multi-author entry that begins with the same name. -->
<!-- If the reference list contains two or more items by the same author(s) in the same year, add a, b, etc. and list them alphabetically by title. -->
<!-- Successive entries by two or more authors when only the first author is the same are alphabetized by co-authors' surnames. -->
<!-- If a reference has more than ten named authors, list only the first seven, followed by 'et al.'. -->
<!-- If a reference has no author or editor, order by title; if a date of publication is impossible to find, use `n.d.' in its place. -->

<!-- The following list shows some sample references prepared in the Taylor & Francis Chicago author-date style. -->

<!-- [@Blair1977; @Alb05] -->
