---
title: |
  On the Scaling of Volatility with Level in Agricultural Commodity Prices and Yields
type: ARTICLE TEMPLATE
author:
  - name: Benjamin Schiek
    affil: a, \dagger
affiliation:
  - num: a
    address: |
      Alliance of Bioversity International and CIAT, Km 17 Recta Cali-Palmira, Valle del Cauca, Colombia, CP 763537
bibliography: interactcadsample.bib
appendix: appendix.tex
abstract: |
  This note documents a strong positive relationship between average prices and price volatility across agricultural commodities using long-run FAO data. The pattern is robust and suggests a simple scaling relationship between level and volatility. We then provide a brief theoretical argument showing that an analogous scaling between yield level and volatility is consistent with basic homogeneity properties of production technology, under which sensitivity to shocks increases with output. The results highlight the importance of scale in shaping volatility and have implications for how risk is conceptualized in agricultural production.
keywords: |
  Sections; lists; figures; tables; mathematics; fonts; references; appendices
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
---

```{r setup, include=FALSE}
library(tidyverse)
library(patchwork)

knitr::opts_chunk$set(echo = TRUE)

# Plot parameters
shapeVec <- c(21:25, 4)
pointSize <- 1.5
smallPointSize <- 1
labelSize <- 2.5
smallLabelSize <- 2
titleSize <- 8
subtitleSize <- 7
legendTextSize <- 7
axisTextSize <- 6
axisTitleSize <- 7
facetTitleSize <- 7

#Folder
thisFolder <- "D:/OneDrive - CGIAR/Documents 1/CIAT 2/Volatility is proportionate to level in agricultural commodity prices/Data/"

# Functions
processDf <- function(thisFilepath, theseElements, theseUnits, isExpPrice = F){
  dfRaw <- read.csv(thisFilepath, stringsAsFactors = F)
  dfRaw <- dfRaw %>% subset(Element %in% theseElements &
                            Area == "World" & Unit %in% theseUnits) %>%
  select(-contains(".")) %>% select(-matches("[NF]$")) %>% rename_with(~ str_remove(., "^Y"))
  gathercols <- colnames(dfRaw)[-c(1:4)]
  df <- dfRaw %>%
  select(-Unit) %>%
  pivot_longer(cols = all_of(gathercols), names_to = "year", values_to = "val")
  # Keep only items with complete time series
  df <- df %>% group_by(Area, Element, Item) %>% mutate(nYrs = length(year))
  #hist(df$nYrs)
  df <- df %>% subset(nYrs == 64)
  # Keep only items with no zero values in their series
  df <- df %>% group_by(Element, Item) %>% mutate(n0 = sum(val == 0)) %>% subset(n0 == 0)
  if(isExpPrice){
  df <- df %>% pivot_wider(names_from = Element, values_from = val) %>%
        mutate(price = `Export value` / `Export quantity`) %>%
        filter(is.finite(price))
  }
  return(df)
}

plotPriceSeries <- function(thisDfPlotTs, plotTitle, yTitle){
  gg <- ggplot(thisDfPlotTs, aes(x = year, y = priceSeries, color = Item, group = Item))
  gg <- gg + geom_line(lwd = 1)
  gg <- gg + labs(y = yTitle,
                title = plotTitle)
  gg <- gg + theme_bw()
  gg <- gg + theme(axis.title.y = element_text(size = axisTitleSize),
                 axis.text = element_text(size = axisTextSize),
                 legend.text = element_text(size = legendTextSize),
                 plot.title = element_text(size = titleSize),
                 axis.title.x = element_blank(),
                 legend.title = element_blank(),
                 legend.position = "bottom")
  return(gg)
}

processRiskRewardDf <- function(thisDf){
  dfRiskReward <- thisDf %>% group_by(Item) %>% summarise(mPrice = mean(priceSeries),
                                          sPrice = sd(diff(priceSeries)[-1])) %>%
  mutate(lmPrice = log(mPrice), lsPrice = log(sPrice)) %>% select(Item, lmPrice, lsPrice)
  dfRiskReward <- dfRiskReward %>%
  mutate(group = case_when(
    # 1. Animal products
    str_detect(Item, regex("Animal|Dairy|Meat|Bovine|Beef|Pig|Poultry|Sheep|Goat|Milk|Cheese|Butter|Egg|Yoghurt|Whey|Fat of|Honey|Beeswax|Tallow|Ghee|Offal|Lard|Fat|Fatty|Cream|Yoghurt", ignore_case = TRUE)) ~ "Animal products",

    # 2. Cereals & starchy roots
    str_detect(Item, regex("Chicory|Taro|Wheat|Fonio|Rice|Maize|Barley|Oats|Sorghum|Millet|Cereal|Buckwheat|Rye|Grain|Bran|Cassava|Potato|Sweet potatoes|Yams|Roots and Tubers|Starch of|Bread|Pastry|pasta", ignore_case = TRUE)) ~ "Cereals & starchy roots",

    # 3. Fruit & vegetables
    str_detect(Item, regex("Beet pulp|Mangoes|Cucumber|Eggplants (aubergines)|Pumpkins|Lemons|Okra|Papayas|Lettuce|Vegetable|Cherries|Apples|Bananas|Oranges|Grapes|Citrus|Berries|Onions|Tomatoes|Vegetables|Fruit|Juice|Melons|Mushrooms|Asparagus|Cabbages|Pineapples|Plantains|Garlic|Spinach|Peaches|Pears|Apricots|Raisins|Artichoke|Avocado|Cauliflower|Chillies|Carrot|Currants|Dates|Figs|Cider|Plums|Persimmons|Tangerines|Quinces", ignore_case = TRUE)) ~ "Fruit & vegetables",

    # 4. Pulses, nuts, oils
    str_detect(Item, regex("Beans|palm fruit|Peas|Lentils|Chick peas|Pulses|Almonds|Cashew|Walnut|Hazelnut|Nuts|Pistachio|Areca|Oil|Soy|Sunflower|Rape|Palm|Coconut|Olive|Linseed|Sesame|Cake of|Groundnut|Mustard|Castor|Margarine|Copra|Karite|seed", ignore_case = TRUE)) ~ "Pulses, nuts, oils",

    # 5. Sugar & stimulants
    str_detect(Item, regex("Anise|Spice|Hop|Sugar|stimulant|Molasses|Coffee|Tea|Cocoa|Maté|Confectionery|Chocolate|Glucose|Lactose|Alcoholic|Wine|Beer|Beverages|Ethyl alcohol|Cinnamon|Cloves|Ginger|Nutmeg|Malt|Pepper|Vanilla", ignore_case = TRUE)) ~ "Sugar, spices, stimulants",

    # 6. Industrial/non-food
    str_detect(Item, regex("Pyrethrum, dried flowers|Ramie, raw or retted|Sisal, raw|Pellets|Cotton|Wool|Silk|Rubber|Tobacco|Hides|Skins|Abaca|Jute|Coir|Flax|Cigarettes|Cigars|Straw|Hair|Waste|Residues|wax|feed|Hay|forage|materials|Textile|Fibre|cat food", ignore_case = TRUE)) ~ "Industrial/feed/non-food",

    # Default for miscellaneous prep or missed items
    TRUE ~ "Other"
  ))
return(dfRiskReward)
}

getSlopeEtc <- function(dfRiskReward){
  dfMod <- dfRiskReward %>% select(lmPrice, lsPrice) %>% as.data.frame()
row.names(dfMod) <- dfRiskReward$Item
mod <- lm(lmPrice ~., dfMod)
#summary(mod)
b <- round(mod$coefficients[1], 2)
m <- round(mod$coefficients[2], 2)
#plot(mod$fitted.values, mod$residuals)
#---
# gg <- ggplot(data.frame(Item = row.names(dfMod), yhat = mod$fitted.values,
#                         resid = mod$residuals),
#              aes(x = yhat, y = resid, label = Item))
# gg <- gg + geom_point() + geom_text()
# gg <- gg + geom_hline(yintercept = c(-1, 1), color = "red")
#---
# ind_outlier <- which(abs(as.numeric(mod$residuals)) >= 0.75)
# excluded_items <- row.names(dfMod)[ind_outlier]
# pct_removed <- 100 * length(ind_outlier) / nrow(dfRiskReward)
# row_names <- row.names(dfMod)[-ind_outlier]
# dfMod <- dfMod[-ind_outlier, ]
# row.names(dfMod) <- row_names
# mod <- lm(lmPrice ~., dfMod)
# #summary(mod)
# b <- round(mod$coefficients[1], 3)
# m <- round(mod$coefficients[2], 3)
dfOut <- as.data.frame(broom::glance(mod))
adjR2 <- round(dfOut$adj.r.squared, 2)
N <- df.residual(mod)
outVec <- c(slope = m, yInt = b, adjR2 = adjR2, N = N)
return(outVec)
}


plotRiskReward <- function(dfPlot, labelItems, thisTitle = NULL, thisXlab = NULL, thisYlab = NULL, m, b, N, adjR2){
  dfPlot$labelThese <- NA
  u <- dfPlot$Item
  dfPlot$labelThese[which(u %in% labelItems)] <- u[which(u %in% labelItems)]
#---
  thisSubtitle <- paste0("N = ", N, " food and non-food agricultural commodities", "\nAdj. R-squared = ", adjR2, ", Slope = ", m, ", Y intercept = ", b)
#---
thisTitle <- paste0(thisTitle,"\n", thisSubtitle)
#names(facetLabel) <- facetKey
#colnames(dfPlot)[2:3] <- c("Logged mean", "Logged stand. dev.")
#---
n <- length(unique(dfPlot$group))
bag_of_colors <- randomcoloR::distinctColorPalette(k = 2 * n)
colorVec <- sample(bag_of_colors, n)
#---
dfPlot$group <- as.factor(dfPlot$group)
names(colorVec) <- levels(dfPlot$group)
names(shapeVec) <- levels(dfPlot$group)
fillScale <- scale_fill_manual(name = "group", values = colorVec)
shapeScale <- scale_shape_manual(name = "group", values = shapeVec)
#---
gg <- ggplot(dfPlot,
             aes(x = lmPrice,
                 y = lsPrice,
                 group = group, fill = group,
                 shape = group,
                 label = labelThese))
gg <- gg + geom_smooth(aes(group = NULL, fill = NULL, shape = NULL), method = lm, se = F)
gg <- gg + geom_point(alpha = 0.6, size = pointSize, color = "black", stroke = 0.5)#,
#fill = "khaki1", pch=21, color="Black")
#gg <- gg + scale_fill_manual(values = color_vec)
# gg <- gg + scale_shape_manual(values = shape_vec)
gg <- gg + shapeScale
gg <- gg + fillScale
gg <- gg + ggrepel::geom_text_repel(size = labelSize,
                           point.padding = 1,
                           box.padding = 1,
                           nudge_x = 1,
                           nudge_y = 1)
gg <- gg + labs(title = thisTitle, x = thisXlab, y = thisYlab)
#gg <- gg + facet_wrap(~Element, labeller = labeller(Element = facetLabel))
#gg <- gg + labs(title = "The ag commodity producer's risk-reward frontier",
                # subtitle = this_subtitle)
gg <- gg + guides(fill = guide_legend(nrow = 1, byrow = T, override.aes = list(linetype = 0)),
                  color = guide_legend(override.aes = list(linetype = 0)))
gg <- gg + theme(legend.position = "bottom",
                 legend.spacing.x = unit(0.25, 'cm'),
                 legend.title = element_blank(),
                 strip.background = element_rect(fill = "white"),
                 strip.text = element_text(size = facetTitleSize),
                 legend.text = element_text(size = legendTextSize),
                 plot.title = element_text(size = titleSize),
                 plot.subtitle = element_text(size = subtitleSize),
                 axis.title = element_text(size = axisTitleSize),
                 axis.text = element_text(size = axisTextSize))
return(gg)
  
}



```

# Introduction

Modern conventional agricultural research for development (AR4D) narratives and agroecological counter-narratives alike agree on the importance of yield stability and price risk in smallholder planting decisions and welfare outcomes [@conway2007doubly; @altieri2015agroecology]. That is to say, it is generally accepted that the primary objectives of AR4D extend beyond genetic gain and smallholder participation in high value commodity markets to also encompass agrononomic and financial stability. However, little consideration has been given to what extent it is possible to pursue the objectives of higher value and stability simultaneously, or whether a stability-reward tradeoff exists, such that the pursuit of more reward (higher yields and price) necessarily entails higher risk (yield and price volatility).

In this note, I present two complementary premises---one empirical and the other theoretical---suggesting that a tradeoff does indeed exist between smallholder value and stability. In other words, smallholder risk and reward are proportional to each other. First, I document a robust empirical regularity across agricultural commodities: higher yielding commodities exhibit higher yield volatility; and higher priced commodities likewise exhibit higher price risk. Second, I present a brief theoretical argument showing that, under the standard assumption of homogeneous production technology, yield sensitivity to input perturbations for any single given commodity necessarily scales with yield level.
<!-- a cursory but rigorous exploration of this question based on FAO data. The result of the exploration indicates a strong empirical correspondence between yield risk and level, and also between price risk and level.  -->
<!-- Likewise, the higher a commodity yield volatility is likewise proportionate to yield level, indicating a risk–reward tradeoff at the commodity level. -->

Taken together, these results suggest that yield and price volatility are not merely exogenous uncertainty, but are structurally inherent to scale itself. AR4D projects aiming to raise yields and increase smallholder participation in high value commodity markets should complement with mechanisms   resilience building 

Strategies to get smallholders to adopt high yielding varieties must complement the technology with resilience building cooperative social mechanisms risk pooling.

```{r Fig12, fig.show = "hold", fig.width = 6, fig.height=7, fig.align="center", fig.cap="\\label{fig:riskReward}Means plotted against standard deviations of (panel A) world export price, (panel B) world producer kilogram price, (panel C) world producer hectare price, and (panel D) yield, for agricultural commodities, 1961-2024. Export price = Deflated export value / export quantity. Producer kg price = FAO Value of Agricultural Production (VAP) series divided / Production. Producer ha price = VAP series divided by area harvested. All data are downloaded from FAO (FAO 2026). Nominal dollar values are deflated to constant 2014-16 USD using the World Bank Manufacturers Unit Value Index (World Bank 2026).", echo = FALSE}

#============================================================================
# Export price series
#============================================================================
thisFile <- "Trade_CropsLivestock_E_All_Data.csv"
thisFilepath <- paste0(thisFolder, thisFile)
theseElements <- c("Export quantity", "Export value")
theseUnits <- c("t", "1000 USD")
df <- processDf(thisFilepath, theseElements, theseUnits, isExpPrice = T)

# To deflate historical USD values, FAO uses the WB MUV series.
# Let's do the same.
# Available on the "Pink Sheet" (annual) https://www.worldbank.org/en/research/commodity-markets
thisFile <- "CMO-Historical-Data-Annual.xlsx"
thisFilepath <- paste0(thisFolder, thisFile)
dfMUV <- readxl::read_xlsx(thisFilepath, sheet = "Annual Indices (Real)")
dfMUV <- dfMUV %>% select(year = 1, MUV = 18) %>% slice(-c(1:8))
dfMUV$MUV <- as.numeric(dfMUV$MUV)
# Rebase MUV to more suitable year dollars (default is 2010 dollars)
# (Rebase to match FAO 2014-2016 constant USD used in FAO VAP data)
muvRebaser <- dfMUV %>%
  filter(year %in% c("2014", "2015", "2016")) %>%
  summarise(muvRebaser = mean(MUV)) %>%
  pull(muvRebaser)
df <- df %>% merge(dfMUV) %>% mutate(priceSeries = price * muvRebaser / MUV)
# # Get historical CPI deflator series (Historical CPI-U) from Bureao of Labor Statistics
# #https://www.bls.gov/cpi/tables/supplemental-files/
# # Use "Index Averages" worksheet (the annual avg over months)
# # Note this deflates to 1982-84 dollars. To rebase to 2024 dollars: CPIrebased = cpiSeries / (cpi2024value / 100)
# thisFile <- "historical-cpi-u-202512.xlsx"
# thisFilepath <- paste0(thisFolder, thisFile)
# dfCPI <- readxl::read_excel(thisFilepath, sheet = "Index Averages")
# dfCPI <- dfCPI %>% select(year = 2, avgCPI = 5) %>% slice(-c(1:5))
# dfCPI$avgCPI <- as.numeric(dfCPI$avgCPI)
# # Rebase CPI to more suitable year dollars (default is 1982-84 dollars)
# # (Rebase to match FAO 2014-2016 constant USD used in VAP data)
# cpiRebaser <- dfCPI %>%
#   filter(year %in% c("2014", "2015", "2016")) %>%
#   summarise(cpiRebaser = mean(avgCPI)) %>%
#   pull(cpiRebaser)
# df <- df %>% merge(dfCPI) %>% mutate(priceSeries = price * cpiRebaser / avgCPI)
#----------------------------------------------------------------------------
# Check stationarity
# library(Kendall)
# df <- df %>% group_by(Item) %>% mutate(MK_kgPrice = as.numeric(MannKendall(priceSeries)[1]))
# hist(df$MK_kgPrice)
# library(fable)
# library(feasts)
# library(tsibble)
# library(features)
# # 1. Convert your df to a tsibble
# df_ts <- df %>%
#   mutate(year = as.integer(year)) %>%
#   as_tsibble(index = year, key = Item)
# # 2. Calculate features
# df_features <- df_ts %>% features(priceSeries, feat_stl)
# # 3. Look at the 'trend_strength' column
# # This will give you a value 0-1 for every commodity
# head(df_features)
#----------------------------------------------------------------------------
# Plot a random sample of deflated commodity world export price series
# num_to_plot <- 5
# randomItems <- sample(unique(df$Item), num_to_plot)
#df$Item[grep("Rice", df$Item)] %>% unique()
plotThese <- c("Wheat", "Rice, milled", "Beans, dry", "Natural Rubber", "Beef and Buffalo Meat, primary", "Sugar Crops Primary")
yTitle <- "World export real price\n(2014-2016 USD / kg)"
plotTitle <- "World export real price series\n(1961-2024) for selected commodities"
dfPlotTs <- df %>% mutate(year = as.integer(year)) %>% subset(Item %in% plotThese)
gg <- plotPriceSeries(thisDfPlotTs = dfPlotTs, plotTitle, yTitle)
ggSampleTsExpPrice <- gg
#----------------------------------------------------------------------------
# Risk reward plot
dfRiskReward <- processRiskRewardDf(df)
dfRiskReward <- dfRiskReward %>% subset(group != "Other" & Item != "Crude Materials nes")
#----------------------------------------------------------------------------
# Get slope and intercept of the risk reward relationship
outVec <- getSlopeEtc(dfRiskReward)
m <- outVec[1]; b <- outVec[2]; adjR2 <- outVec[3]; N <- outVec[4];
#----------------------------------------------------------------------------
#dfPlot$Item[grep("Maize", dfPlot$Item, ignore.case = T)]
labelItems <- c("Maize (corn)", "Flour of maize",
  #"Wheat", "Flour, wheat",
  #"Starch, cassava",
                "Coffee, green", "Coffee, decaffeinated or roasted", #"Bananas",
                #"Plantains",
                #"Sweet potatoes",
                #"Potatoes",
                "Cocoa beans",
                #"Wool, greasy", 
                #"Beans, dry",
                "Raw silk (not thrown)",
                #"Apples",
  "Vanilla, raw",
                "Soya beans", "Soya bean oil", "Natural rubber", "Rice",
                "Cocoa butter, fat and oil")
#---
thisTitle <- "World export real price (2014-16 USD/kg)\nMean and standard deviation over years 1961-2024"
thisXlab <- "Ln(mean real price, 1961-2024)"; thisYlab <- "Ln(stand. dev. of differenced real price, 1962-2024)"
ggRiskRewardExpPrice <- plotRiskReward(dfRiskReward, labelItems, thisTitle, thisXlab, thisYlab, m, b, N, adjR2)
#============================================================================
# Farmgate price series
#============================================================================
thisFile <- "Production_Crops_Livestock_E_All_Data.csv"
thisFilepath <- paste0(thisFolder, thisFile)
theseElements <- c("Production", "Area harvested", "Yield"); theseUnits <- c("t", "ha", "kg/ha")
dfAreaProd <- processDf(thisFilepath, theseElements, theseUnits, isExpPrice = F) %>%
  spread(Element, val) %>% select(Item, year, Production, `Area harvested`, Yield)
thisFile <- "Value_of_Production_E_All_Data.csv"
thisFilepath <- paste0(thisFolder, thisFile)
theseElements <- "Gross Production Value (constant 2014-2016 thousand US$)"; theseUnits <- "1000 USD"
dfVAP <- processDf(thisFilepath, theseElements, theseUnits, isExpPrice = F) %>%
  select(Item, year, val) %>% rename(VAP = val); dfVAP$Element <- NULL
df <- dfAreaProd %>% merge(dfVAP, by = c("Item", "year")) %>%
  mutate(haPrice = 1000 * VAP / `Area harvested`, kgPrice = VAP / Production,
         year = as.integer(year))
#----------------------------------------------------------------------------
# Check stationarity
# library(Kendall)
# # Example for one item:
# df <- df %>% group_by(Item) %>% mutate(MK_kgPrice = as.numeric(MannKendall(kgPrice)[1]), MK_haPrice = as.numeric(MannKendall(haPrice)[1]))
# hist(df$MK_kgPrice)
# hist(df$MK_haPrice)
# x <- MannKendall(df$kgPrice[df$Item == "Abaca, manila hemp, raw"])
#----------------------------------------------------------------------------
# Plot a random sample of deflated commodity world price series
#df$Item[grep("Soya beans", df$Item, ignore.case = T)] %>% unique()
#"Coffee, green", "Wheat", "Soya beans",
plotThese <- c("Sugar Crops Primary", "Walnuts, in shell",
               "Pistachios, in shell", "Beeswax",
              "Raspberries", "Sesame seed")
# num_to_plot <- 15
# randomItems <- sample(unique(df$Item), num_to_plot)
# plotThese <- randomItems
yTitle <-"World producer real price\n(2014-2016 USD / kg)"
plotTitle <- "World producer real price series\n(1961-2024) for selected commodities (constant 2014-2016 USD)"
dfPlotTs <- df %>% subset(Item %in% plotThese) %>% rename(priceSeries = kgPrice)
gg <- plotPriceSeries(dfPlotTs, plotTitle, yTitle)
ggSampleTsProdPrice <- gg
yTitle <-"World producer real price per hectare\n(2014-2016 USD / hectare)"
plotTitle <- "World producer real production value per hectare series\n(1961-2024) for selected commodities (constant 2014-2016 USD)"
dfPlotTs <- df %>% subset(Item %in% plotThese) %>% rename(priceSeries = haPrice)
gg <- plotPriceSeries(dfPlotTs, plotTitle, yTitle)
ggSampleTsProdHaPrice <- gg
#----------------------------------------------------------------------------
# Risk reward plot
dfKg <- df %>% rename(priceSeries = kgPrice)
dfRiskReward <- processRiskRewardDf(dfKg)
dfRiskReward <- dfRiskReward %>% subset(group != "Other")
#----------------------------------------------------------------------------
# Get slope and intercept of the risk reward relationship
outVec <- getSlopeEtc(dfRiskReward)
m <- outVec[1]; b <- outVec[2]; adjR2 <- outVec[3]; N <- outVec[4];
#----------------------------------------------------------------------------
labelItems <- c("Maize (corn)",
                "Coffee, green",
                "Bananas",
                #"Plantains",
                #"Sweet potatoes",
                #"Potatoes",
                "Cocoa beans",
                #"Wool, greasy", 
                "Beans, dry",
                "Silk-worm cocoons suitable for reeling",
                #"Apples",
                "Vanilla, raw",
                "Soya beans", "Natural rubber in primary forms", "Rice")
thisTitle <- "World producer real price (2014-16 USD/kg)\nMean and standard deviation over years 1961-2024"
thisXlab <- "Ln(mean world real producer price/kg, 1961-2024)"; thisYlab <- "Ln(stand. dev. of differenced world real producer price/kg, 1962-2024)"
ggRiskRewardProdKgPrice <- plotRiskReward(dfRiskReward, labelItems, thisTitle, thisXlab, thisYlab, m, b, N, adjR2)
#----------------------------------------------------------------------------
dfYd <- df %>% rename(priceSeries = Yield)
dfRiskReward <- processRiskRewardDf(dfYd)
dfRiskReward <- dfRiskReward %>% subset(group != "Other")
#----------------------------------------------------------------------------
# Get slope and intercept of the risk reward relationship
outVec <- getSlopeEtc(dfRiskReward)
m <- outVec[1]; b <- outVec[2]; adjR2 <- outVec[3]; N <- outVec[4];
#----------------------------------------------------------------------------
thisTitle <- "World yield (2014-16 USD/ha)\nMean and standard deviation over years 1961-2024"
thisXlab <- "Ln(mean yield, 1961-2024)"; thisYlab <- "Ln(stand. dev. of differenced yield, 1962-2024)"
ggRiskRewardYd <- plotRiskReward(dfRiskReward, labelItems, thisTitle, thisXlab, thisYlab, m, b, N, adjR2)
#----------------------------------------------------------------------------
dfHa <- df %>% rename(priceSeries = haPrice)
dfRiskReward <- processRiskRewardDf(dfHa)
dfRiskReward <- dfRiskReward %>% subset(group != "Other")
#----------------------------------------------------------------------------
# Get slope and intercept of the risk reward relationship
outVec <- getSlopeEtc(dfRiskReward)
m <- outVec[1]; b <- outVec[2]; adjR2 <- outVec[3]; N <- outVec[4];
#----------------------------------------------------------------------------
thisTitle <- "World producer real price per hectare (2014-16 USD)\nMean and standard deviation over years 1961-2024"
thisXlab <- "Ln(mean real price/ha, 1961-2024)"; thisYlab <- "Ln(stand. dev. of differenced real price/ha, 1962-2024)"
ggRiskRewardProdHaPrice <- plotRiskReward(dfRiskReward, labelItems, thisTitle, thisXlab, thisYlab, m, b, N, adjR2)
#===============================================================================
# Put together the 2x2 display of panels
# 1. Arrange the plots in a 2x2 grid
# The "/" operator stacks plots vertically, while "|" or "+" aligns them horizontally
combined_plot <- (ggRiskRewardExpPrice | ggRiskRewardProdKgPrice) / (ggRiskRewardProdHaPrice | ggRiskRewardYd)

# 2. Add labels and a global theme
combined_plot + 
  plot_annotation(
    tag_levels = 'A'#,                      # Use uppercase letters A, B, C, D
    #title = "Empirical evidence of a reward-stability tradeoff in commodity prices and yields",
    #subtitle = "Analysis of yield stability vs. initial infection rate",
    #caption = "Source: Pathogen Model v1.2"
  ) & 
  theme(plot.tag = element_text(face = 'bold')) # Apply bold to all tags at once

```

# Empirical evidence of an agricultural commodity reward-stability tradeoff

To assess the relationship between agricultural risk and reward, the mean and standard deviation of global yield and price series over 1961-2024 are plotted against each other in logs (Figure \ref{fig:riskReward}). Only commodities with a complete, non-zero time series are included. The comparison of means and standard deviations of time series is meaningful insofar as the series are stationary. The standard deviations in each plot are taken over the first differenced time series to guarantee stationarity. In the price series (panels A-C), dollar values are deflated using the World Bank Manufacturers Unit Value Index [@WB_MUV_2026], such that the means are generally (but not always) stationary. The means of yield series, on the other hand, should be interpreted with great caution, since yield series generally exhibit a trend.

Taylor...

# Theoretical basis for a yield level-stability tradeoff

In the cultivation of any single commodity, there is a fundamental theoretical reason to expect a tradeoff between yield level and yield stability. Under the standard assumption of yield $y$ as a homogeneous function of inputs $\mathbf{x}$, such that, by Euler's theorem,

\begin{equation}
\sum_{i=1}^n x_i \frac{\partial y}{\partial x_i}
= h y(\mathbf{x})
\end{equation}

where $h$ is the degree of homogeneity, or in vector notation,

\begin{equation}
\mathbf{x}' \nabla y(\mathbf{x})
= h y(\mathbf{x}).
\end{equation}

then, differentiating,

\begin{equation}
\nabla \left( \mathbf{x}' \nabla y \right)
= \nabla (h y).
\end{equation}

and applying the product rule,

\begin{equation}
\nabla y + H(\mathbf{x}) \mathbf{x}
= h \nabla y,
\end{equation}

where $H(\mathbf{x})$ is the Hessian matrix.

Rearranging,

\begin{equation}
H(\mathbf{x}) \mathbf{x}
= (h - 1)\nabla y.
\end{equation}

and premultiplying by $\mathbf{x}'$ gives the quadratic form

\begin{equation}
\mathbf{x}' H(\mathbf{x}) \mathbf{x}
= (h - 1) \mathbf{x}' \nabla y.
\end{equation}

Which, invoking Euler’s theorem once more, reduces to

\begin{equation}
\mathbf{x}' H(\mathbf{x}) \mathbf{x}
= h(h - 1) y(\mathbf{x}).
\end{equation}

Thus

\begin{equation}
\mathbf{x}' H(\mathbf{x}) \mathbf{x}
\propto y(\mathbf{x}).
\label{eq:volatPropToLev}
\end{equation}

The curvature measure $\mathbf{x}' H(\mathbf{x}) \mathbf{x}$ indicates yield sensitivity to small changes in inputs---in other words, the fragility of yield if inputs (like rainfall) are slightly off their optimal values---and may thus be interpreted as a measure of yield volatility. Equation \ref{eq:volatPropToLev} thus indicates that yield level is proportional to yield volatility.

[for decreasing returns to scale]

## Interpretation

The curvature measure:

\begin{equation}

\end{equation}

scales proportionally with yield.

Since the Hessian captures local sensitivity of output to perturbations in inputs, this implies:

> Yield sensitivity (volatility) scales with yield level under homogeneous production.

For decreasing returns:

\begin{equation}
0 < h < 1 \quad \Rightarrow \quad h(h - 1) < 0.
\end{equation}

---

## Summary Result

\begin{equation}
\boxed{
\mathbf{x}' H(\mathbf{x}) \mathbf{x}
= h(h - 1) y(\mathbf{x})
}
\end{equation}












Conclusion:
crop, animal, food, non-food, raw, processed, individual or aggregated
smallholder risk mitigation, AIS [@sandlin2014agricultural]




















with implications for productivity measurement, varietal choice, and smallholder risk management.

In this note, I present empirical evidence of a tradeoff between participation in high value commodity markets and price volatility. 

higher yields and market exposure, on the one hand, and yield stability and price volatility, on the other. An assessment of the relation

Conventional Green Revolution narratives of agricultural research for development (AR4D) have evolved in the early 21st century to appreciate the importance of yield stability and price risk in smallholder planting decisions and welfare outcomes [Conway, Derek?, AIS]. This evolution brings the conventional narrative into partial alignment with the agroecological counter-narrative [], which is largely premised upon a sharp critique of the conventional narrative's perceived myopic fixation on genetic gain and participation in high value commodity markets. 


Conventional Green Revolution narratives of agricultural research for development (AR4D)---traditionally focused on genetic gain and access to high value commodity markets [e.g. Pingali]---have evolved in the early 21st century to appreciate the importance of yield stability and price risk in smallholder planting decisions and welfare outcomes [Conway, Derek?, AIS]. This evolution brings the conventional narrative into partial alignment with agroecological counter-narratives premised upon a longstanding critique of conventional research's perceived myopic fixation on genetic gain []. Both narrative and counter-narrative now agree, in principle, that stability and resilience are important objectives to pursue simultaneously with the pursuit of higher yields and greater participation in high value commodity markets. However, both stop short of assessing the relationship that may exist between these objectives. In particular, they stop short of considering the potential tradeoffs (synergies) that may exist, such that progress towards one objective nullifies (enhances) progress towards the other.



While the new-found consensus is welcome, narrative and counternarrative alike recognize the importance... However, little ... Or does a tradeoff exist such that progress along one dimension effectively nullifies progress along another? 

narrative, sometimes called the "Doubly Green Revolution" [Conway] [[or the "new agricultural paradigm"?? [Byerlee]]], is in violent agreement with the (otherwise sharply critical) agroecology counternarrative on certain key issues [Altieri, Mockshell, Holly guy]. In particular, the scope of AR4D narratives has expanded beyond genetic gain to include stability (financial and agronomic) and social-ecological resilience [Conway].  However, little consideration has been given to the tradeoffs or synergies that might exist between these dimensions.

Most agricultural development narratives now recognize, at least in principle, the importance of stable farm yield and income in addition to increased output and market value, especially on smallholder farms [AEI, SAI (conway), Mockshell, ]. The 





* AEI vs SAI [] ...efforts at synthesis [Jonathan]. Some of the original vision for SAI included the key elements of AEI [Conway]. Farmers risk averse []. While this work draws attention to the importance of other objectives beyond yield, it stops short of assessing the relationship between objectives.    stability and level. In particular, does a relation exist between higher production and farm return volatility? If so, is the relation proportionate or inverse? This is important to know is the relation a tradeoff?

* In this note, I present empirical evidence and a theoretical argument suggesting a risk-reward tradeoff.



Existing efforts note that stability/resilience is important in addition to the pursuit of genetic gain and into high value commodity markets; but this work stops short of characterizing or formalizing the relationships between these dimensions. To better characterize these relations, here I present empirical evidence of a tradeoff between commodity market risk and reward. I also present a brief theoretical argument indicating yield volatility scales with yield level under the standard assumption of homogenous production technology.

Existing work emphasizes that yield stability and resilience matter alongside genetic gain and market upgrading, particularly for risk-averse smallholders. However, this literature largely stops short of characterizing how volatility scales with level, either empirically in markets or formally in production.



There is no inherent statistical relationship between the mean and standard deviation of normally distributed variables. The scaling relationship documented here is therefore not a consequence of normality, but an empirical regularity across commodities. Similar mean–variance scaling relationships have been documented in other domains (e.g. Taylor’s law), where they are interpreted as reflecting structural or technological features rather than probabilistic identities.

This note documents a strong positive relationship between average prices and price volatility across agricultural commodities using long-run FAO data.

The purpose of the production-side argument is not to explain price behavior, but to clarify a general point: volatility need not reflect larger shocks, but can arise mechanically from scale-dependent sensitivity.









In order to assist authors in the process of preparing a manuscript for a journal, the Taylor & Francis '\textsf{Interact}' layout style has been implemented as a \LaTeXe\ class file based on the `article` document class. A \textsc{Bib}\TeX\ bibliography style file and a sample bibliography are also provided in order to assist with the formatting of your references.

Commands that differ from or are provided in addition to standard \LaTeXe\ are described in this document, which is _not_ a substitute for a \LaTeXe\ tutorial.

The `interactcadsample.tex` file can be used as a template for a manuscript by cutting, pasting, inserting and deleting text as appropriate, using the preamble and the \LaTeX\ environments provided (e.g.\ `\begin{abstract}`, `\begin{keywords}`).

## The \textsf{Interact} class file

The `interact` class file preserves the standard \LaTeXe\ interface such that any document that can be produced using `article.cls` can also be produced with minimal alteration using the `interact` class file as described in this document.

If your article is accepted for publication it will be typeset as the journal requires in Minion Pro and/or Myriad Pro. Since most authors will not have these fonts installed, the page make-up is liable to alter slightly with the change of font. Also, the `interact` class file produces only single-column format, which is preferred for peer review and will be converted to two-column format by the typesetter if necessary during preparation of the proofs. Please therefore do not try to match the typeset format exactly, but use the standard \LaTeX\ fonts instead and ignore details such as slightly long lines of text or figures/tables not appearing in exact synchronization with their citations in the text: these details will be dealt with by the typesetter. Similarly, it is unnecessary to spend time addressing warnings in the log file -- if your .tex file compiles to produce a PDF document that correctly shows how you wish your paper to appear, such warnings will not prevent your source files being imported into the typesetter's program.

## Submission of manuscripts prepared using _\LaTeX_

Manuscripts for possible publication should be submitted to the Editors for review as directed in the journal's Instructions for Authors, and in accordance with any technical instructions provided in the journal's ScholarOne Manuscripts or Editorial Manager site. Your \LaTeX\ source file(s), the class file and any graphics files will be required in addition to the final PDF version when final, revised versions of accepted manuscripts are submitted.

Please ensure that any author-defined macros used in your article are gathered together in the preamble of your .tex file, i.e.\ before the `\begin{document}` command. Note that if serious problems are encountered in the coding of a document (missing author-defined macros, for example), the typesetter may resort to rekeying it.

# Using the `interact` class file

For convenience, simply copy the `interact.cls` file into the same directory as your manuscript files (you do not need to install it in your \TeX\ distribution). In order to use the `interact` document class, replace the command `\documentclass{article}` at the beginning of your document with the command `\documentclass{interact}`.

The following document-class options should _not_ be used with the `interact` class file:

* `10pt`, `11pt`, `12pt` -- unavailable;
* `oneside`, `twoside` -- not necessary, `oneside` is the default;
* `leqno`, `titlepage` -- should not be used;
* `twocolumn` -- should not be used;
* `onecolumn` -- not necessary as it is the default style.

To prepare a manuscript for a journal that is printed in A4 (two column) format, use the `largeformat` document-class option provided by `interact.cls`; otherwise the class file produces pages sized for B5 (single column) format by default. The `geometry` package should not be used to make any further adjustments to the page dimensions.

<!-- If your manuscript has supplementary content you can prepare this using the `suppldata` document-class option, which will suppress the `article history' date. This option must _not_ be used on any primary content. -->

# Additional features of the `interact` class file

## Title, authors' names and affiliations, abstracts and article types

## Abbreviations

A list of abbreviations may be included if required, enclosed within an `abbreviations` environment, i.e.\ `\begin{abbreviations}`\ldots`\end{abbreviations}`, immediately following the `abstract` environment.


## Keywords

A list of keywords may be included if required, enclosed within a `keywords` environment, i.e.\ `\begin{keywords}`\ldots`\end{keywords}`. Additional keywords in other languages (preceded by a translation of the word `keywords`) may also be included within the `keywords` environment if required.


## Subject classification codes

AMS, JEL or PACS classification codes may be included if required. The `interact` class file provides an `amscode` environment, i.e.\ `\begin{amscode}`\ldots`\end{amscode}`, a `jelcode` environment, i.e.\ `\begin{jelcode}`\ldots`\end{jelcode}`, and a `pacscode` environment, i.e.\ `\begin{pacscode}`\ldots`\end{pacscode}` to assist with this.


## Additional footnotes to the title or authors' names

<!-- The `\thanks` command may be used to create additional footnotes to the title or authors' names if required. Footnote symbols for this purpose should be used in the order -->
<!-- $^\ast$~(coded as `$^\ast$`), $\dagger$~(`$\dagger$`), $\ddagger$~(`$\ddagger$`), $\S$~(`$\S$`), $\P$~(`$\P$`), $\|$~(`$\|$`), -->
<!-- $\dagger\dagger$~(`$\dagger\dagger$`), $\ddagger\ddagger$~(`$\ddagger\ddagger$`), $\S\S$~(`$\S\S$`), $\P\P$~(`$\P\P$`). -->

Note that any `footnote`s to the main text will automatically be assigned the superscript symbols 1, 2, 3, etc. by the class file.^[If preferred, the `endnotes` package may be used to set the notes at the end of your text, before the bibliography. The symbols will be changed to match the style of the journal if necessary by the typesetter.]

# Embedding R code

## Code chunks


```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE, out.width='80%'}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

# Some guidelines for using the standard features of \LaTeX

## Sections

The \textsf{Interact} layout style allows for five levels of section heading, all of which are provided in the `interact` class file using the standard \LaTeX\ commands `\section`, `\subsection`, `\subsubsection`, `\paragraph` and `\subparagraph`. Numbering will be automatically generated for all these headings by default.

## Lists

Numbered lists are produced using the `enumerate` environment, which will number each list item with arabic numerals by default. For example,

1. first item
2. second item
3. third item

Alternative numbering styles can be achieved by inserting an optional argument in square brackets to each `item`, e.g.\ `\item[(i)] first item`\, to create a list numbered with roman numerals at level one.

Bulleted lists are produced using the `itemize` environment. For example,

* First bulleted item
* Second bulleted item
* Third bulleted item

## Figures

```{r pressure-plot, fig.show='hide'}
plot(pressure)
```

The `interact` class file will deal with positioning your figures in the same way as standard \LaTeX. It should not normally be necessary to use the optional `[htb]` location specifiers of the `figure` environment in your manuscript; you may, however, find the `[p]` placement option or the `endfloat` package useful if a journal insists on the need to separate figures from the text.

Figure captions appear below the figures themselves, therefore the `\caption` command should appear after the body of the figure. For example, Figure~\ref{sample-figure} with caption and sub-captions is produced using the following commands:
```
\begin{figure}
\centering
\subfigure[An example of an individual figure sub-caption.]{%
\resizebox*{5cm}{!}{\includegraphics{path/to/fig}}}\hspace{5pt}
\subfigure[A slightly shorter sub-caption.]{%
\resizebox*{5cm}{!}{\includegraphics{path/to/fig}}}
\caption{Example of a two-part figure with individual sub-captions
 showing that captions are flush left and justified if greater
 than one line of text.} \label{sample-figure}
\end{figure}
```
\begin{figure}
\centering
\subfigure[An example of an individual figure sub-caption.]{%
\resizebox*{5cm}{!}{\includegraphics{`r knitr::fig_chunk('pressure-plot', 'pdf')`}}}\hspace{5pt}
\subfigure[A slightly shorter sub-caption.]{%
\resizebox*{5cm}{!}{\includegraphics{`r knitr::fig_chunk('pressure-plot', 'pdf')`}}}
\caption{Example of a two-part figure with individual sub-captions
 showing that captions are flush left and justified if greater
 than one line of text.} \label{sample-figure}
\end{figure}

To ensure that figures are correctly numbered automatically, the `\label` command should be included just after the `\caption` command, or in its argument.

The `\subfigure` command requires `subfigure.sty`, which is called in the preamble of the `interacttfssample.tex` file (to allow your choice of an alternative package if preferred) and included in the \textsf{Interact} \LaTeX\ bundle for convenience. Please supply any additional figure macros used with your article in the preamble of your .tex file.

The source files of any figures will be required when the final, revised version of a manuscript is submitted. Authors should ensure that these are suitable (in terms of lettering size, etc.) for the reductions they envisage.

The `epstopdf` package can be used to incorporate encapsulated PostScript (.eps) illustrations when using PDF\LaTeX, etc. Please provide the original .eps source files rather than the generated PDF images of those illustrations for production purposes.


## Tables

The `interact` class file will deal with positioning your tables in the same way as standard \LaTeX. It should not normally be necessary to use the optional `[htb]` location specifiers of the `table` environment in your manuscript; you may, however, find the `[p]` placement option or the `endfloat` package useful if a journal insists on the need to separate tables from the text.

The `tabular` environment can be used as shown to create tables with single horizontal rules at the head, foot and elsewhere as appropriate. The captions appear above the tables in the \textsf{Interact} style, therefore the `\tbl` command should be used before the body of the table. For example, Table~\ref{sample-table} is produced using the following commands:
\begin{table}
\tbl{Example of a table showing that its caption is as wide as
 the table itself and justified.}
{\begin{tabular}{lcccccc} \toprule
 & \multicolumn{2}{l}{Type} \\ \cmidrule{2-7}
 Class & One & Two & Three & Four & Five & Six \\ \midrule
 Alpha\textsuperscript{a} & A1 & A2 & A3 & A4 & A5 & A6 \\
 Beta & B2 & B2 & B3 & B4 & B5 & B6 \\
 Gamma & C2 & C2 & C3 & C4 & C5 & C6 \\ \bottomrule
\end{tabular}}
\tabnote{\textsuperscript{a}This footnote shows how to include
 footnotes to a table if required.}
\label{sample-table}
\end{table}
```
\begin{table}
\tbl{Example of a table showing that its caption is as wide as
 the table itself and justified.}
{\begin{tabular}{lcccccc} \toprule
 & \multicolumn{2}{l}{Type} \\ \cmidrule{2-7}
 Class & One & Two & Three & Four & Five & Six \\ \midrule
 Alpha\textsuperscript{a} & A1 & A2 & A3 & A4 & A5 & A6 \\
 Beta & B2 & B2 & B3 & B4 & B5 & B6 \\
 Gamma & C2 & C2 & C3 & C4 & C5 & C6 \\ \bottomrule
\end{tabular}}
\tabnote{\textsuperscript{a}This footnote shows how to include
 footnotes to a table if required.}
\label{sample-table}
\end{table}
```

To ensure that tables are correctly numbered automatically, the `\label` command should be included just before `\end{table}`.

The `\toprule`, `\midrule`, `\bottomrule` and `\cmidrule` commands are those used by `booktabs.sty`, which is called by the `interact` class file and included in the \textsf{Interact} \LaTeX\ bundle for convenience. Tables produced using the standard commands of the `tabular` environment are also compatible with the `interact` class file.


## Landscape pages

If a figure or table is too wide to fit the page it will need to be rotated, along with its caption, through 90$^{\circ}$ anticlockwise. Landscape figures and tables can be produced using the `rotating` package, which is called by the `interact` class file. The following commands (for example) can be used to produce such pages.
```
\setcounter{figure}{1}
\begin{sidewaysfigure}
\centerline{\epsfbox{figname.eps}}
\caption{Example landscape figure caption.}
\label{landfig}
\end{sidewaysfigure}
```
```
\setcounter{table}{1}
\begin{sidewaystable}
 \tbl{Example landscape table caption.}
  {\begin{tabular}{@{}llllcll}
    .
    .
    .
  \end{tabular}}\label{landtab}
\end{sidewaystable}
```
Before any such float environment, use the `\setcounter` command as above to fix the numbering of the caption (the value of the counter being the number given to the preceding figure or table). Subsequent captions will then be automatically renumbered accordingly. The `\epsfbox` command requires `epsfig.sty`, which is called by the `interact` class file and is also included in the \textsf{Interact} \LaTeX\ bundle for convenience.

Please note that if the `endfloat` package is used, one or both of the commands
```
\DeclareDelayedFloatFlavor{sidewaysfigure}{figure}
\DeclareDelayedFloatFlavor{sidewaystable}{table}
```
will need to be included in the preamble of your .tex file, after the `endfloat` package is loaded, in order to process any landscape figures and/or tables correctly.


## Theorem-like structures

A predefined `proof` environment is provided by the `amsthm` package (which is called by the `interact` class file), as follows:
\begin{proof}
More recent algorithms for solving the semidefinite programming relaxation are particularly efficient, because they explore the structure of the MAX-CUT problem.
\end{proof}
\noindent This was produced by simply typing:
```
\begin{proof}
More recent algorithms for solving the semidefinite programming
relaxation are particularly efficient, because they explore the
structure of the MAX-CUT problem.
\end{proof}
```
Other theorem-like environments (theorem, definition, remark, etc.) need to be defined as required, e.g.\ using `\newtheorem{theorem}{Theorem}` in the preamble of your .tex file (see the preamble of `interactcadsample.tex` for more examples). You can define the numbering scheme for these structures however suits your article best. Please note that the format of the text in these environments may be changed if necessary to match the style of individual journals by the typesetter during preparation of the proofs.


## Mathematics

### Displayed mathematics

The `interact` class file will set displayed mathematical formulas centred on the page without equation numbers if you use the `displaymath` environment or the equivalent `\[...\]` construction. For example, the equation
\[
 \hat{\theta}_{w_i} = \hat{\theta}(s(t,\mathcal{U}_{w_i}))
\]
was typeset using the commands
```
\[
 \hat{\theta}_{w_i} = \hat{\theta}(s(t,\mathcal{U}_{w_i}))
\]
```

For those of your equations that you wish to be automatically numbered sequentially throughout the text for future reference, use the `equation` environment, e.g.
\begin{equation}
 \hat{\theta}_{w_i} = \hat{\theta}(s(t,\mathcal{U}_{w_i}))
\end{equation}
was typeset using the commands
```
\begin{equation}
 \hat{\theta}_{w_i} = \hat{\theta}(s(t,\mathcal{U}_{w_i}))
\end{equation}
```

Part numbers for sets of equations may be generated using the `subequations` environment, e.g.
\begin{subequations} \label{subeqnexample}
\begin{equation}
     \varepsilon \rho w_{tt}(s,t) = N[w_{s}(s,t),w_{st}(s,t)]_{s},
     \label{subeqnparta}
\end{equation}
\begin{equation}
     w_{tt}(1,t)+N[w_{s}(1,t),w_{st}(1,t)] = 0,
     \label{subeqnpartb}
\end{equation}
\end{subequations}
which was typeset using the commands
```
\begin{subequations} \label{subeqnexample}
\begin{equation}
     \varepsilon \rho w_{tt}(s,t) = N[w_{s}(s,t),w_{st}(s,t)]_{s},
     \label{subeqnparta}
\end{equation}
\begin{equation}
     w_{tt}(1,t)+N[w_{s}(1,t),w_{st}(1,t)] = 0,   \label{subeqnpartb}
\end{equation}
\end{subequations}
```
<!-- This is made possible by the `amsmath` package, which is called by the class file. If you put a `\label` just after the `\begin{subequations}` command, references can be made to the collection of equations, i.e.\ `(\ref{subeqnexample})' in the example above. Or, as the example also shows, you can label and refer to each equation individually -- i.e.\ `(\ref{subeqnparta})' and `(\ref{subeqnpartb})'. -->

Displayed mathematics should be given end-of-line punctuation appropriate to the running text sentence of which it forms a part, if required.

### Math fonts

#### Superscripts and subscripts
Superscripts and subscripts will automatically come out in the correct size in a math environment (i.e.\ enclosed within `\(...\)` or `$...$` commands in running text, or within `\[...\]` or the `equation` environment for displayed equations). Sub/superscripts that are physical variables should be italic, whereas those that are labels should be roman (e.g.\ $C_p$, $T_\mathrm{eff}$). If the subscripts or superscripts need to be other than italic, they must be coded individually.

#### Upright Greek characters and the upright partial derivative sign
Upright lowercase Greek characters can be obtained by inserting the letter `u` in the control code for the character, e.g.\ `\umu` and `\upi` produce $\umu$ (used, for example, in the symbol for the unit microns -- $\umu\mathrm{m}$) and $\upi$ (the ratio of the circumference of a circle to its diameter). Similarly, the control code for the upright partial derivative $\upartial$ is `\upartial`. Bold lowercase as well as uppercase Greek characters can be obtained by `{\bm \gamma}`, for example, which gives ${\bm \gamma}$, and `{\bm \Gamma}`, which gives ${\bm \Gamma}$.


# Acknowledgement(s) {-}

An unnumbered section, e.g.\ `\section*{Acknowledgements}`, may be used for thanks, etc.\ if required and included _in the non-anonymous version_ before any Notes or References.

# Disclosure statement {-}

An unnumbered section, e.g.\ `\section*{Disclosure statement}`, may be used to declare any potential conflict of interest and included _in the non-anonymous version_ before any Notes or References, after any Acknowledgements and before any Funding information.

# Funding {-}

An unnumbered section, e.g.\ `\section*{Funding}`, may be used for grant details, etc.\ if required and included _in the non-anonymous version_ before any Notes or References.

# Notes on contributor(s) {-}

An unnumbered section, e.g.\ `\section*{Notes on contributors}`, may be included _in the non-anonymous version_ if required. A photograph may be added if requested.

# Nomenclature/Notation {-}

An unnumbered section, e.g.\ `\section*{Nomenclature}` (or `\section*{Notation}`), may be included if required, before any Notes or References.

# Notes {-}

An unnumbered `Notes` section may be included before the References (if using the `endnotes` package, use the command `\theendnotes` where the notes are to appear, instead of creating a `\section*`).

# References

## References cited in the text

## The list of references

References should be listed at the end of the main text in alphabetical order by authors' surnames, then chronologically (earliest first). If references have the same author(s), editor(s), etc., arrange by year of publication, with undated works at the end.
A single-author entry precedes a multi-author entry that begins with the same name.
If the reference list contains two or more items by the same author(s) in the same year, add a, b, etc. and list them alphabetically by title.
Successive entries by two or more authors when only the first author is the same are alphabetized by co-authors' surnames.
If a reference has more than ten named authors, list only the first seven, followed by 'et al.'.
If a reference has no author or editor, order by title; if a date of publication is impossible to find, use `n.d.' in its place.

The following list shows some sample references prepared in the Taylor & Francis Chicago author-date style.

[@Blair1977; @Alb05]
